---
title: "Relative Humidity"
author: "Larz von Huene"
date: '2022-06-06'
output: html_document
---

This rmd calculates relative humidity using WMO annex 4.B and analyzes its trends annually, seasonally, and monthly. The relative humidity is used to determine the presence of fog, and then fog frequency is calculated and analyzed.


#Load Packages
```{r, message=FALSE}
library(tidyverse)
library(lubridate)
library(trend)
library(patchwork)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)
```

# Load in Data
```{r, show_col_types = FALSE, message = FALSE}
synoptic <- read_csv("Data/b16 synoptic data 1935-2021.csv") %>% 
  dplyr::select(-c(precipitation, precipitation_trace))
pres_weather <- read_csv("Data/b16 hourly present weather split by code 1935-2021.csv")

# Adding 2022 for synoptic
bulb <- read_csv("Data/2022_synoptic_wet_dry_bulb.csv") %>% 
  mutate(date = as.POSIXct(date, format = "%m/%d/%Y %H:%M", tz = "ETC"))
pressure <- read_csv("Data/2022_RH_pressure.csv") %>%
  dplyr::select(-relative_humidity)  # takes out rh measurement from observatory so we can calculate all rh the same way 

synoptic_2022 <- full_join(bulb, pressure, by = "date") 
  
synoptic <- rbind(synoptic, synoptic_2022) # adding 2022
```


# Clean data
```{r}
#convert temperature into °C
synoptic <- synoptic %>% 
  mutate(dry_bulb = as.numeric(dry_bulb), 
         wet_bulb = as.numeric(wet_bulb), 
         pressure_mb = as.numeric(pressure_mb)) %>% # switch from chr to dbl
  mutate(dry_bulb = (dry_bulb-32)*(5/9),
         wet_bulb = (wet_bulb-32)*(5/9))

# Quality control
synoptic <- synoptic %>% 
  filter(wet_bulb != "NULL" &  pressure_mb != "NULL") %>%  # removes all null values (from wet_bulb, dry_bulb, and pressure_mb)
  filter(wet_bulb <= dry_bulb) %>% # ensures that wet bulb is less than or equal to dry bulb for the sake of the RH equation
  filter(dry_bulb-wet_bulb < 10)  # no gaps larger than 10 degrees between wet and dry)

```



# Calculate Relative Humidity and other metrics (Saturation vapor pressure, dewpoint, mixing ratio)
Calculations from WMO annex 4.B, mixing ratio equation from American Meterology Society glossary
Methods for calculation are based on those from Seidel (2007)
```{r}
## Liquid calculations

#rh
rh_liquid <- function(t, tw, p){
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p) 
  
  sat_vap_pure = 6.112*(exp((17.62*t)/(243.12+t)))
  sat_vap_pure_from_tw = 6.112*(exp((17.62*tw)/(243.12+tw)))
  
  sat_vap_moist = f_p * sat_vap_pure
  
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)
  
  return(100*vap_p/sat_vap_moist)
}

#saturation vapor pressure
sat_vap_liquid <- function(t) {
  return(6.112*(exp((17.62*t)/(243.12+t))))
}

#vapour pressure
vap <- function(t, tw, p){
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p) 
  
  sat_vap_pure_from_tw = 6.112*(exp((17.62*tw)/(243.12+tw)))
  
  return(sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw))
}


#dewpoint
dewpoint <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  
  sat_vap_pure = 6.112*(exp((17.62*t)/(243.12+t)))
  sat_vap_pure_from_tw = 6.112*(exp((17.62*tw)/(243.12+tw)))
  
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)
  
  return((243.12*log(vap_p/(6.112*f_p)))/(17.62-log(vap_p/(6.112*f_p))))
}

#mixing ratio
mixing_ratio <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  sat_vap_pure_from_tw = 6.112*(exp((17.62*tw)/(243.12+tw)))
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)
  
  return((0.622*vap_p)/(p-vap_p))
}


liquid <- synoptic %>% 
  filter(dry_bulb >= 0) %>% # Seidel's threshold
  mutate(RH = rh_liquid(dry_bulb, wet_bulb, pressure_mb),
         sat_vap = sat_vap_liquid(dry_bulb),
         dewpoint = dewpoint(dry_bulb, wet_bulb, pressure_mb),
         vap = vap(dry_bulb, wet_bulb, pressure_mb),
         mixing_ratio = mixing_ratio(dry_bulb, wet_bulb, pressure_mb))


## Ice calculations

#rh
rh_ice <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  
  sat_vap_pure = 6.112*(exp((22.46*t)/(272.62+t)))
  sat_vap_pure_from_tw = 6.112*(exp((22.46*tw)/(272.62+tw)))
  
  sat_vap_moist = f_p * sat_vap_pure
  
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw) # Seidel (2007) used psychometric formula for liquid water for both water and ice
  
  return(100*vap_p/sat_vap_moist)
}

#saturation vapour pressure
sat_vap_ice <- function(t) {
  return(6.112*(exp((22.46*t)/(272.62+t))))
}

#vapour pressure
vap <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  
  sat_vap_pure_from_tw = 6.112*(exp((22.46*tw)/(272.62+tw)))
  
  return(sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)) # Seidel (2007) used psychometric formula for liquid water for both water and ice
}

#dewpoint
dewpoint <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  
  sat_vap_pure = 6.112*(exp((22.46*t)/(272.62+t)))
  sat_vap_pure_from_tw = 6.112*(exp((22.46*tw)/(272.62+tw)))
  
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)
  
  return((243.12*log(vap_p/(6.112*f_p)))/(17.62-log(vap_p/(6.112*f_p))))
}

#mixing ratio
mixing_ratio <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  sat_vap_pure_from_tw = 6.112*(exp((22.46*tw)/(272.62+tw)))
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)
  
  return((0.622*vap_p)/(p-vap_p))
}

ice <- synoptic %>% 
  filter(dry_bulb < 0) %>% # Seidel's threshold
  mutate(RH = rh_ice(dry_bulb, wet_bulb, pressure_mb),
         sat_vap = sat_vap_ice(dry_bulb),
         dewpoint = dewpoint(dry_bulb, wet_bulb, pressure_mb), 
         vap = vap(dry_bulb, wet_bulb, pressure_mb),
         mixing_ratio = mixing_ratio(dry_bulb, wet_bulb, pressure_mb))

# Combining
rh <- rbind(liquid, ice)



# Quality control
summary(rh$RH) # 3 NAs
summary(rh$sat_vap)
summary(rh$dewpoint) # 290 NA's
summary(rh$vap)
summary(rh$mixing_ratio) # 3 NA's

nrow(rh[rh$RH<0,]) # 290 negative rh
nrow(rh[rh$vap<0,]) # 290 negative vap

rh <- rh %>% 
  na.omit() %>% 
  filter(RH >= 0) # filter out negative RH (and vap p by association)

# Add seasons 
# seasons by month: so winter is 12 and 1-2, spring is 3-5, summer is 6-8, fall is 9-11
rh <- rh %>% 
  na.omit() %>% 
  mutate(year = year(date), 
         month = month(date), 
         year = as.numeric(year), 
         month = as.numeric(month),
         year_for_winter = ifelse(month == 12, year + 1, year), # this is to correct for month 12 being in previous year, so that winter can be grouped together in season_summary (below) 
         season = case_when(
           month == 12 ~ "Winter",
           month == 1 ~ "Winter",
           month == 2 ~ "Winter",
           month == 3 ~ "Spring",
           month == 4 ~ "Spring",
           month == 5 ~ "Spring",
           month == 6 ~ "Summer",
           month == 7 ~ "Summer",
           month == 8 ~ "Summer",
           month == 9 ~ "Fall",
           month == 10 ~ "Fall",
           month == 11 ~ "Fall",
           TRUE ~ NA_character_)) %>% 
  mutate(tod = case_when(
           hour(date) == 1 ~ "Night",
           hour(date) == 19 ~ "Night",
           hour(date) == 7 ~ "Day",
           hour(date) == 13 ~ "Day",
           TRUE ~ NA_character_))


# Export dataset 
write.csv(rh, file = "Data/rh_from_humidity_analysis_rmd.csv")

rh <- rh %>% 
  filter(year(date) >= 1941)
```



# Averaging
```{r}
# Getting averages first with daily, then monthly, then seasonal
season_summary <- rh %>% 
  group_by(date(date), month, season, year_for_winter) %>% 
  summarize(daily_average_RH = mean(RH),
            daily_average_sat_vap = mean(sat_vap)) %>% # first get daily averages
  group_by(month, season, year_for_winter) %>% 
  summarize(monthly_average_RH = mean(daily_average_RH),
            monthly_average_sat_vap = mean(daily_average_sat_vap)) %>% # get monthly avg from daily avg
  group_by(season, year_for_winter) %>% 
  summarize(average_RH = mean(monthly_average_RH),
            average_sat_vap = mean(monthly_average_sat_vap)) # get seasonal avg from monthly avg
  

# comparing with Seidel 2007
season_summary %>% 
  filter(year_for_winter <= 2004) %>% 
  group_by(season) %>% 
  summarize(avg = mean(average_RH))


# Clear day differences in seasons
pdf(file = "Figures/rh_density_plot.pdf", width = 8, height = 5)
rh %>% 
  filter(RH < 100) %>% 
  mutate(season = factor(season, c("Winter", "Spring", "Summer", "Fall"))) %>% 
  ggplot(aes(x = RH, fill = season, color = season)) + geom_density(alpha = 0.5) + 
  theme_classic() +
  labs(x = "Relative Humidity (%)", y = "Density")
dev.off()

rh %>% 
  filter(RH < 100) %>% 
  mutate(season = factor(season, c("Winter", "Spring", "Summer", "Fall"))) %>% 
  ggplot(aes(x = dewpoint, fill = season, color = season)) + geom_density(alpha = 0.5) + 
  theme_classic()

rh %>% 
  filter(RH < 100) %>% 
  mutate(season = factor(season, c("Winter", "Spring", "Summer", "Fall"))) %>% 
  ggplot(aes(x = dewpoint, fill = season, color = season)) + geom_density(alpha = 0.5) + 
  theme_classic()
```


# Seasonal comparisons

```{r}
## Sen Slopes for seasons

# Sen's slope
trends <- season_summary %>% 
  group_by(season) %>% 
  summarize(trend = sens.slope(average_RH, conf.level = 0.95)[["estimates"]],
            p = sens.slope(average_RH, conf.level = 0.95)[["p.value"]])

## VISUALS

# all seasons on one graph
season_plot <- season_summary %>% 
  ggplot(aes(year_for_winter, average_RH, color = season)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  facet_wrap(~season) +
  labs(title = "Average Relative Humidities by Season", subtitle = "1935-2022", x = "Year", y = "Relative Humidity (%)", caption = "Note: Winter graph groups winter months consecutively (e.g. December of 1996, January of 1997, and February of 1997)") 

ggsave(season_plot, file = "Figures/humidities_seasons.jpeg", width = 8, height = 5, dpi = 30)
```



# Month plot: summary
```{r}
# adding names to months
month_rh <- month_rh %>% 
  filter(year >= 1941) %>% # for consistency with wind data
   mutate(month_name = case_when(month == 1 ~ "January", 
                                month == 2 ~ "February",
                                month == 3 ~ "March",
                                month == 4 ~ "April",
                                month == 5 ~ "May",
                                month == 6 ~ "June",
                                month == 7 ~ "July",
                                month == 8 ~ "August",
                                month == 9 ~ "September",
                                month == 10 ~ "October",
                                month == 11 ~ "November",
                                month == 12 ~ "December")) 
month_rh$month_name <- factor(month_rh$month_name, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
  


# RH by month
pdf(file = "Figures/avg_humidity_over_year.pdf", width = 8, height = 5)
ggplot(month_rh, aes(as.factor(month_name), average_RH)) + 
  geom_boxplot(fill = "grey", outlier.shape = 1) + 
  theme_classic() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  labs(x = "Month", y = "Relative Humidity (%)") +
  ggtitle("Monthly Average Relative Humidities 1941-2022")
dev.off()
```


# Improving visual
```{r}
month_rh %>%
  ggplot(aes(year, average_RH)) +
  facet_wrap(~month_name) +
  geom_point() +
  geom_smooth(method = lm, se = F) +
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal() 

#trends 
month_rh %>% 
  group_by(month_name) %>% 
  summarize(trend = sens.slope(average_RH, conf.level = 0.95)[["estimates"]],
            p = sens.slope(average_RH, conf.level = 0.95)[["p.value"]])
```

# Overall RH over years 
```{r}
# getting averages 
year_rh <- rh %>% 
  group_by(date(date), month, season, year) %>% 
  summarize(daily_average_RH = mean(RH),
            daily_average_sat_vap = mean(sat_vap)) %>% # first get daily averages
  group_by(month, season, year) %>% 
  summarize(monthly_average_RH = mean(daily_average_RH),
            monthly_average_sat_vap = mean(daily_average_sat_vap)) %>% # get monthly avg from daily avg
  group_by(year) %>% 
  summarize(avg_rh = mean(monthly_average_RH),
            avg_sat_vap = mean(monthly_average_sat_vap)) # get yearly avg from monthly avg

# overall average 1941 - 2022
year_rh %>% 
  summarize(avg_rh = mean(avg_rh))


pdf(file = "Figures/rh_annual_1941.pdf", width = 8, height = 5)
year.sen <- sens.slope(year_rh$avg_rh, conf.level = 0.95)
ggplot(year_rh, aes(year, avg_rh)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal() +
  ylim(70, 100) +
  xlim(1941, 2022) + # select only years after 1941 for continuity with wind data
  labs(x = "Year", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidities 1941-2022") + 
  annotate("text", x=2015, y=97, label = paste("p-value =", round((year.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=99, label = paste("Sen's Slope =", round((year.sen[["estimates"]][["Sen's slope"]]), digits = 3)))
dev.off()

# add temp to same graph
rh_all_fig <- ggplot(year_rh, aes(year, avg_rh)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal() +
  ylim(70, 100) +
  xlim(1941, 2022) + # select only years after 1941 for continuity with wind data
  labs(x = "Year", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidities 1941-2022") + 
  annotate("text", x=2015, y=97, label = paste("p-value =", round((year.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=99, label = paste("Sen's Slope =", round((year.sen[["estimates"]][["Sen's slope"]]), digits = 3)))

temp <- rh %>% 
  group_by(date(date), month, season, year) %>% 
  summarize(daily_average_temp = mean(dry_bulb)) %>% # first get daily averages
  group_by(month, season, year) %>% 
  summarize(monthly_average_temp = mean(daily_average_temp)) %>% # get monthly avg from daily avg
  group_by(year) %>% 
  summarize(avg_temp = mean(monthly_average_temp)) # get yearly avg from monthly avg

temp_fig <- ggplot(temp, aes(year, avg_temp)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal()

sat_vap_fig <- ggplot(year_rh, aes(year, avg_sat_vap)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal()

print(temp_fig/rh_all_fig/sat_vap_fig)
```



# In the clear vs. in the clouds
```{r}
# adding fog and clear class
rh <- rh %>% 
  mutate(Class = ifelse(RH == 100, "Fog", "Clear")) %>% 
  dplyr::select(date, RH, year, month, year_for_winter, season, Class)

# filter for clear 
clear <- rh %>% 
  filter(Class == "Clear") 

# average RH each year
clear_summary <- clear %>% 
  group_by(month, year) %>% 
  summarize(avg_rh = mean(RH)) %>% 
  group_by(year) %>% 
  summarize(avg_rh = mean(avg_rh))

# average in the clear 1941 - 2022 
clear_summary %>% 
  summarize(mean(avg_rh))

# analysis (Sen's Slope)
clear.sen <- sens.slope(clear_summary$avg_rh, conf.level = 0.95)
#View(clear.sen)
clear.sen[["p.value"]]
clear.sen[["conf.int"]]



# clear vs all visual
rh_clear_all <- ggplot(year_rh, aes(x = year, y = avg_rh)) + 
  geom_line(data = year_rh, aes(x = year, y = avg_rh), color = "grey50") + geom_smooth(data = year_rh, aes(x = year, y = avg_rh), method = "lm", se = F, color = "grey70", size = 0.5) + 
  geom_line(data = clear_summary, aes(x = year, y = avg_rh), color = "skyblue3", lty = "longdash") + geom_smooth(data = clear_summary, aes(x = year, y = avg_rh), method = "lm", se = F, color = "skyblue2", size = 0.5) +
  theme_minimal() + 
  labs(x = "Year", y = "Relative Humidity (%)", title = "Relative humidity overall and in the clear", subtitle = "1941-2022" )

ggsave(rh_clear_all, file = "Figures/rh_clear_all_plot.jpeg", width = 8, height = 5, dpi = 30)
```


# Fog frequency
```{r}
# Fog frequency by season: all days 1941-2022
ff_seasonal <- rh %>% 
  group_by(season, Class) %>% 
  summarise(n_class = n()) %>% # get number of fog and clear days
  group_by(season) %>% 
  summarise(n_season = sum(n_class), n_class = n_class, Class = Class) %>%  # get total number of days in season
  summarise(ff = n_class/n_season, Class = Class)

#what statistical test will look for significant differences between seasons?

# overall fog frequency
ff_overall <- rh %>% 
  group_by(Class) %>% 
  summarise(n_class = n()) %>% 
  mutate(n_total = sum(n_class)) %>% 
  summarise(ff = n_class/n_total, Class = Class)

# visual: seasonal          SHOULD I INCORPORATE OVERALL HERE?
pdf(file = "Figures/ff_season.pdf", width = 8, height = 5)
ggplot(ff_seasonal, aes(x = season, y = ff, fill = Class)) + 
  geom_col() +
  scale_fill_manual(values =c("skyblue", "darkgrey")) +
  theme_minimal() +
  labs(title = "Frequency of days spent in the clear and in the fog", subtitle = "1935-2022", x = "Season", y = "")
dev.off()
  
  
# trends

# annual
ff_annual <- rh %>% 
  group_by(Class, year) %>% 
  summarise(n_class = n()) %>% 
  group_by(year) %>% 
  mutate(n_total = sum(n_class)) %>% 
  group_by(Class, year) %>% 
  summarise(ff = n_class/n_total, Class = Class) %>% 
  filter(Class == "Fog")

# visual
ggplot(ff_annual, aes(x= year, y = ff)) + geom_point() + geom_smooth(method = "sen", se = F)

ff_sen <- sens.slope(ff_annual$ff, conf.level = 0.95)


# seasonal
ff_seasonal <- rh %>% 
  group_by(Class, year, season) %>% 
  summarise(n_class = n()) %>% 
  group_by(year, season) %>% 
  mutate(n_total = sum(n_class)) %>% 
  group_by(Class, year, season) %>% 
  summarise(ff = n_class/n_total, Class = Class) %>% 
  filter(Class == "Fog")




ff_annual_trend <- ff_annual %>% 
  group_by(year) %>% 
  summarize(annual_trend = sens.slope(ff_annual$ff, conf.level = 0.95)[["estimates"]],
            p = sens.slope(ff_annual$ff, conf.level = 0.95)[["p.value"]]) 
  

ff_seasonal_trend <- ff_seasonal %>% 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>% 
  group_by(year, season) %>% 
  summarise(avg_ff = mean(ff)) %>% 
  group_by(season) %>% 
  summarize(annual_trend = sens.slope(avg_ff, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_ff, conf.level = 0.95)[["p.value"]]) 


# combining for table
ff_annual_trends_all <- bind_rows(ff_annual_trend, ff_seasonal_trend)  %>% 
  mutate(decadal_trend = annual_trend*10*100) %>% #switches trend to decadal percentage
  mutate(" " = ifelse(is.na(season), "Annual", as.character(season)), 
         decadal_trend = format(round(decadal_trend, 3), nsmall = 3), # round trend to 3 decimal places
         "Decadal trend (%)" = ifelse(p <= 0.05, str_c(decadal_trend, "*"), decadal_trend), # add asterisks
         "Decadal trend (%)" = ifelse(p <= 0.1, str_c(decadal_trend, "\u2020"), decadal_trend), # add dagger
         "p-value" = p) %>% 
  dplyr::select(" ", "Decadal trend (%)") %>% 
  distinct() %>%   # removes rows with repeat values
  as.data.frame(html_table()) 


ff_table <- ff_annual_trends_all %>% 
  kbl(caption = "Decadal Trends of Fog Frequency from 1941-2022") %>%
  footnote(general = c("\u2020 indicates 90% confidence", "*indicates 95% confidence", "Sen's Slope analysis used to calculate annual fog frequency trends")) %>% 
  kable_classic(full_width = F, html_font = "Cambria")  

save_kable(ff_table, file = "Figures/ff_table.pdf")

```



