---
title: "Relative Humidity"
author: "Larz von Huene"
date: '2022-06-06'
output: html_document
---
#Relative humidity calculations and analysis 
# Packages and set working directory
```{r}
library(tidyverse)
#library(psychrolib)
library(lubridate)
library(trend)
library(patchwork)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)
# setwd("//wdc-fileserver/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends")
```

# Load in Data
```{r, show_col_types = FALSE}
synoptic <- read_csv("Data/b16 synoptic data 1935-2021.csv")
pres_weather <- read_csv("Data/b16 hourly present weather split by code 1935-2021.csv")
```
# Clean data
```{r}
synoptic <- synoptic %>% 
  filter(wet_bulb != "NULL" &  pressure_mb != "NULL") %>%  # removes all null values (from wet_bulb, dry_bulb, and pressure_mb)
  mutate(dry_bulb = as.numeric(dry_bulb), 
         wet_bulb = as.numeric(wet_bulb), 
         pressure_mb = as.numeric(pressure_mb), 
         precipitation = as.numeric(precipitation)) %>% # switch from chr to dbl
  filter(wet_bulb <= dry_bulb) %>% # ensures that wet bulb is less than or equal to dry bulb for the sake of the RH equation
  filter(dry_bulb-wet_bulb < 10) %>%  # no gaps larger than 10 degrees between wet and dry)
  filter() # NEED TO UPDATE: no 20 degree increases between readings

  
```



# Calculate Relative Humidity 
```{r}

# Calculations from WMO annex 4.B

# split data into ice and water (based on Seidel's distinction)
synoptic_rh_liquid <- synoptic %>% 
  filter(wet_bulb >= 32) 
synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

## calculations for water
# saturation vapor pressure
f_p_water <- 1.0016+3.15*(10^-6)*(synoptic_rh_liquid$pressure_mb)-0.074*(synoptic_rh_liquid$pressure_mb)^-1

sat_vap_p_water_pure <- 6.112^(17.62*(synoptic_rh_liquid$dry_bulb)/(243.12+(synoptic_rh_liquid$dry_bulb))) # saturation vapor pressure for pure water
sat_vap_p_water_moist <- f_p_water*sat_vap_p_water_pure  # saturation vapor pressure for moist water
  
# psychometric formulas for the Assmann psychrometer
water_vap_p <- (6.112^(17.62*(synoptic_rh_liquid$wet_bulb)/(243.12+(synoptic_rh_liquid$wet_bulb))))*f_p_water - 6.53*(10^-4)*(1+0.000944*(synoptic_rh_liquid$wet_bulb))*(synoptic_rh_liquid$pressure_mb)*(synoptic_rh_liquid$dry_bulb - synoptic_rh_liquid$wet_bulb)

# Relative humidity
rh_liquid <- 100*water_vap_p/sat_vap_p_water_moist



# liquid tibble
synoptic_rh_liquid <- synoptic_rh_liquid %>% 
  mutate(RH = rh_liquid) 
synoptic_rh_liquid

# count how many negative values in "synoptic_RH_liquid"
nrow(synoptic_rh_liquid[synoptic_rh_liquid$RH<0,])
  # 0 negative out of 52,560 rows









## calculations for ice

synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

f_p_ice <- 1.0016+(3.15*(10^-6))*(synoptic_rh_ice$pressure_mb)-0.074/synoptic_rh_ice$pressure_mb

sat_vap_p_ice_pure_t <- 6.112*(exp(22.46*(synoptic_rh_ice$dry_bulb)/(272.62+(synoptic_rh_ice$dry_bulb))))
sat_vap_p_ice_moist_t <- sat_vap_p_ice_pure_t*f_p_ice

sat_vap_p_ice_pure_tw <- 6.112*(exp(22.46*(synoptic_rh_ice$wet_bulb)/(272.62+(synoptic_rh_ice$wet_bulb))))
sat_vap_p_ice_moist_tw <- sat_vap_p_ice_pure_tw*f_p_ice

ice_vap_p <- sat_vap_p_ice_pure_t - 5.75*(10^-4)*(1+0.000944*(synoptic_rh_ice$wet_bulb))*(synoptic_rh_ice$pressure_mb)*(synoptic_rh_ice$dry_bulb - synoptic_rh_ice$wet_bulb)

frost_pt <- (272.62*(log(ice_vap_p/6.112*f_p_ice)))/(22.46-(log(ice_vap_p/6.112*f_p_ice)))

sat_vap_p_from_frost <- 6.112*(exp(22.46*(frost_pt)/(272.62+(frost_pt))))
vap_p_from_frost <- sat_vap_p_from_frost - 5.75*(10^-4)*(1+0.000944*(synoptic_rh_ice$wet_bulb))*(synoptic_rh_ice$pressure_mb)*(synoptic_rh_ice$dry_bulb - synoptic_rh_ice$wet_bulb)

rh_ice <- 100*(vap_p_from_frost/sat_vap_p_from_frost)

# ice tibble
synoptic_rh_ice <- synoptic_rh_ice %>% 
  mutate(RH = rh_ice) 
synoptic_rh_ice



# count how many negative values in "synoptic_RH_ice"
nrow(synoptic_rh_ice[synoptic_rh_ice$RH<0,])
  # 73 negative out of 70,874 rows

# percent with negative RH in "synoptic_RH_ice"
73/70874 # 0.1%
```



# Join together liquid and ice
```{r}
rh <- rbind(synoptic_rh_liquid, synoptic_rh_ice) 

# count how many negative values in "synoptic_RH_ice"
nrow(rh[rh$RH<0,])
  # 73 negative out of 70,874 rows

# percent with negative RH in "synoptic_RH_ice"
73/70874 # 0.1%

# filter out negative rh
rh <- rh %>% 
  filter(RH >= 0)
```


# Separate by seasons
```{r}
# seasons by month: so winter is 12 and 1-2, spring is 3-5, summer is 6-8, fall is 9-11
season_rh <- rh %>% 
  mutate(year = year(date), 
         month = month(date), 
         year = as.numeric(year), 
         month = as.numeric(month),
         year_for_winter = ifelse(month == 12, year + 1, year), # this is to correct for month 12 being in previous year, so that winter can be grouped together in season_summary (below) 
         season = case_when(
           month == 12 ~ "Winter",
           month == 1 ~ "Winter",
           month == 2 ~ "Winter",
           month == 3 ~ "Spring",
           month == 4 ~ "Spring",
           month == 5 ~ "Spring",
           month == 6 ~ "Summer",
           month == 7 ~ "Summer",
           month == 8 ~ "Summer",
           month == 9 ~ "Fall",
           month == 10 ~ "Fall",
           month == 11 ~ "Fall",
           TRUE ~ NA_character_))


season_summary <- season_rh %>% 
  group_by(season, year_for_winter) %>% 
  summarize(average_RH = mean(RH)) 

```

# Seasonal comparisons
```{r}
## Linear regression of seasonal data for all season 1935 - 2021, using seasonal averages

# fall
fall_rh <- season_summary %>% 
  filter(season == "Fall") # isolate fall data

fall.lm = lm(average_RH ~ year_for_winter, data = fall_rh)
summary(fall.lm)
# p-value  = 0.7422
# adjusted r-squared = -0.01047


# spring
spring_rh <- season_summary %>% 
  filter(season == "Spring") # isolate spring data

spring.lm = lm(average_RH ~ year_for_winter, data = spring_rh)
summary(spring.lm)
# p-value  = 0.02176**
# adjusted r-squared = 0.04935

# summer
summer_rh <- season_summary %>% 
  filter(season == "Summer") # isolate summer data

summer.lm = lm(average_RH ~ year_for_winter, data = summer_rh)
summary(summer.lm)
# p-value  = 0.9739
# adjusted r-squared = -0.01175

# winter
winter_rh <- season_summary %>% 
  filter(season == "Winter") # isolate winter data

winter.lm = lm(average_RH ~ year_for_winter, data = winter_rh)
summary(winter.lm)
# p-value  = 0.6936
# adjusted r-squared = -0.009793



## Sen Slopes for seasons

# Fall
fall.sen <- sens.slope(fall_rh$average_RH, conf.level = 0.95)
#View(fall.sen)
fall.sen[["p.value"]] # 0.8373378

# Winter
winter.sen <- sens.slope(winter_rh$average_RH, conf.level = 0.95)
#View(winter.sen)
winter.sen[["p.value"]] # 0.7429325

# Spring
spring.sen <- sens.slope(spring_rh$average_RH, conf.level = 0.95)
#View(spring.sen)
spring.sen[["p.value"]] # 0.01375496**

# Summer
summer.sen <- sens.slope(summer_rh$average_RH, conf.level = 0.95)
#View(summer.sen)
summer.sen[["p.value"]] # 0.8776295


## VISUALS

# all seasons on one graph
season_summary %>% 
  ggplot(aes(year_for_winter, average_RH, color = season)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  facet_wrap(~season) +
  labs(title = "Average Relative Humidities by Season", subtitle = "1935-2021", x = "Year", y = "Relative Humidity (%)", caption = "Note: Winter graph groups winter months consecutively (e.g. December of 1996, January of 1997, and February of 1997)") 

# spring
spring_rh %>% 
  ggplot(aes(year_for_winter, average_RH)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE, color = "grey") +
  theme_minimal() +
  labs(title = "Average Relative Humidities in Spring", subtitle = "1935-2021", x = "Year", y = "Relative Humidity (%)") 

# winter
winter_rh %>% 
  ggplot(aes(year_for_winter, average_RH)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE, color = "grey") +
  theme_minimal() +
  labs(title = "Average Relative Humidities in Winter", subtitle = "1935-2021", x = "Year", y = "Relative Humidity (%)") 
```

Spring is showing significant trends, but with very small slope

# Monthly comparisons, focusing on November 
```{r}
rh <- rh %>% 
  mutate(month = month(date), year = year(date))

month_rh <- rh %>% 
  group_by(month, year) %>% 
  summarize(average_RH = mean(RH))

# overview of all months
ggplot(month_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm) + facet_wrap(~month)


# RH by month
ggplot(month_rh, aes(as.factor(month), average_RH)) + 
  geom_boxplot() + 
  theme_bw() +
  labs(x = "Month", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidity by Month 1935-2022")

# November sen slopes
nov_rh <- month_rh %>% 
  filter(month == 11) 

ggplot(nov_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm)

nov.lm = lm(average_RH ~ year, data = nov_rh)
summary(nov.lm)
# p-value  = 0.13
# adjusted r-squared = 0.01531

nov.sen <- sens.slope(nov_rh$average_RH, conf.level = 0.95)
View(nov.sen)
nov.sen[["p.value"]] # 0.3222527

# June
june_rh <- month_rh %>% 
  filter(month == 6) 

ggplot(june_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm)

june.lm = lm(average_RH ~ year, data = june_rh)
summary(june.lm)
# p-value: 0.3115
# adjusted r-squared 0.000425

# September
sept_rh <- month_rh %>% 
  filter(month == 9) 

ggplot(sept_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm)

sept.lm = lm(average_RH ~ year, data = sept_rh)
summary(sept.lm)
# p-value: 0.09771
# adjusted r-squared: 0.02055

sept.sen <- sens.slope(sept_rh$average_RH, conf.level = 0.95)
View(sept.sen)
sept.sen[["p.value"]] # 0.07970815


# checking with Seidel 2007
rh %>% 
  filter(year(date) <= 2004) %>% 
  summarize(avg = mean(RH))

rh %>% 
  filter(year(date) <= 2004) %>% 
  filter(month(date) == 6 | month(date) == 7 | month(date) == 8) %>% # for JJA
  summarize(avg = mean(RH))

```

```{r}
# adding names to months
month_rh <- month_rh %>% 
  filter(year >= 1941) %>% # for consistency with wind data
   mutate(month_name = case_when(month == 1 ~ "January", 
                                month == 2 ~ "February",
                                month == 3 ~ "March",
                                month == 4 ~ "April",
                                month == 5 ~ "May",
                                month == 6 ~ "June",
                                month == 7 ~ "July",
                                month == 8 ~ "August",
                                month == 9 ~ "September",
                                month == 10 ~ "October",
                                month == 11 ~ "November",
                                month == 12 ~ "December")) 
month_rh$month_name <- factor(month_rh$month_name, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
  


# RH by month
ggplot(month_rh, aes(as.factor(month_name), average_RH)) + 
  geom_boxplot(fill = "grey", outlier.shape = 1) + 
  theme_classic() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  labs(x = "Month", y = "Relative Humidity (%)") +
  ggtitle("Monthly Average Relative Humidities 1941-2021")
```


# Improving visual

```{r}
# JANUARY
jan_rh <- month_rh %>% 
  filter(month == 1) 

jan <- ggplot(jan_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("January") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

jan.sen <- sens.slope(jan_rh$average_RH, conf.level = 0.95)
#View(jan.sen)
jan.sen[["p.value"]] # 0.8603164






# FEBRUARY
feb_rh <- month_rh %>% 
  filter(month == 2) 

feb <- ggplot(feb_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("February") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

feb.sen <- sens.slope(feb_rh$average_RH, conf.level = 0.95)
#View(feb.sen)
feb.sen[["p.value"]] # 0.8603164







# MARCH
mar_rh <- month_rh %>% 
  filter(month == 3) 

mar <- ggplot(mar_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("March") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()


mar.sen <- sens.slope(mar_rh$average_RH, conf.level = 0.95)
#View(mar.sen)
mar.sen[["p.value"]] # 0.1726373










# APRIL
apr_rh <- month_rh %>% 
  filter(month == 4) 

apr <- ggplot(apr_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("April") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()


apr.sen <- sens.slope(apr_rh$average_RH, conf.level = 0.95)
#View(apr.sen)
apr.sen[["p.value"]] # 0.07360794







# MAY
may_rh <- month_rh %>% 
  filter(month == 5) 

may <- ggplot(may_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("May") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()


may.sen <- sens.slope(may_rh$average_RH, conf.level = 0.95)
#View(may.sen)
may.sen[["p.value"]] # 0.1385819








# JUNE
jun_rh <- month_rh %>% 
  filter(month == 6) 

jun <- ggplot(jun_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("June") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

jun.sen <- sens.slope(jun_rh$average_RH, conf.level = 0.95)
#View(jun.sen)
jun.sen[["p.value"]] # 0.4157207





# JULY
jul_rh <- month_rh %>% 
  filter(month == 7) 

jul <- ggplot(jul_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("July") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

jul.sen <- sens.slope(jul_rh$average_RH, conf.level = 0.95)
#View(jul.sen)
jul.sen[["p.value"]] # 0.9357175








# AUGUST
aug_rh <- month_rh %>% 
  filter(month == 8) 

aug <- ggplot(aug_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("August") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

aug.sen <- sens.slope(aug_rh$average_RH, conf.level = 0.95)
#View(aug.sen)
aug.sen[["p.value"]] # 0.2650709







# SEPTEMEBR
sep_rh <- month_rh %>% 
  filter(month == 9) 

sep <- ggplot(sep_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("September") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

sep.sen <- sens.slope(sep_rh$average_RH, conf.level = 0.95)
#View(sep.sen)
sep.sen[["p.value"]] # 0.07970815









# OCTOBER
oct_rh <- month_rh %>% 
  filter(month == 10) 

oct <- ggplot(oct_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("October") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

oct.sen <- sens.slope(oct_rh$average_RH, conf.level = 0.95)
#View(oct.sen)
oct.sen[["p.value"]] # 0.1548998












# NOVEMBER
nov_rh <- month_rh %>% 
  filter(month == 11) 

nov <- ggplot(nov_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("November") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

nov.sen <- sens.slope(nov_rh$average_RH, conf.level = 0.95)
#View(nov.sen)
nov.sen[["p.value"]] # 0.3222527












# DECEMBER
dec_rh <- month_rh %>% 
  filter(month == 12) 

dec <- ggplot(dec_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("December") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

dec.sen <- sens.slope(dec_rh$average_RH, conf.level = 0.95)
#View(dec.sen)
dec.sen[["p.value"]] # 0.9473864


jan + feb + mar + apr + may + jun + jul + aug + sep + oct + nov + dec
```

# Overall RH over years (not split by month)
```{r}
year_rh <- rh %>% 
  group_by(year) %>% 
  summarize(avg_rh = mean(avg_rh))


ggplot(year_rh, aes(year, avg_rh)) +
  geom_point() +
  theme_minimal() +
  ylim(60, 100) +
  xlim(1941, 2021) + # select only years after 1941 for continuity with wind data
  labs(x = "Year", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidities 1941-2021")

year.sen <- sens.slope(year_rh$avg_rh, conf.level = 0.95)
#View(year.sen)
year.sen[["p.value"]]
```



# In the clear vs. in the clouds
```{r}
# adding fog and clear class
season_rh <- season_rh %>% 
  mutate(Class = ifelse(dry_bulb == wet_bulb, "Fog", "Clear")) %>% 
  dplyr::select(date, RH, year, month, year_for_winter, season, Class)

# filter for clear 
clear <- season_rh %>% 
  filter(Class == "Clear") 

# average RH each year
clear_summary <- clear %>% 
  group_by(year) %>% 
  summarize(avg_rh = mean(RH))

# visual: dotplot
clear_summary %>% 
  ggplot(aes(x = year, y = avg_rh)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey") +
  theme_minimal() +
  labs(title = "Average Relative Humidity when in the Clear", subtitle = "1935-2021", x = "Year", y = "Average Relative Humidity (%)")

# analysis (Sen's Slope)
clear.sen <- sens.slope(clear_summary$avg_rh, conf.level = 0.95)
#View(clear.sen)
clear.sen[["p.value"]]
clear.sen[["conf.int"]]
```


# Fog frequency
```{r}
# Fog frequency by season: all days 1935-2021
ff_seasonal <- season_rh %>% 
  group_by(season, Class) %>% 
  summarise(n_class = n()) %>% # get number of fog and clear days
  group_by(season) %>% 
  summarise(n_season = sum(n_class), n_class = n_class, Class = Class) %>%  # get total number of days in season
  summarise(ff = n_class/n_season, Class = Class)

#what statistical test will look for significant differences between seasons?

# overall fog frequency
ff_overall <- season_rh %>% 
  group_by(Class) %>% 
  summarise(n_class = n()) %>% 
  mutate(n_total = sum(n_class)) %>% 
  summarise(ff = n_class/n_total, Class = Class)

# visual: seasonal          SHOULD I INCORPORATE OVERALL HERE?
ggplot(ff_seasonal, aes(x = season, y = ff, fill = Class)) + 
  geom_col() +
  scale_fill_manual(values =c("skyblue", "darkgrey")) +
  theme_minimal() +
  labs(title = "Frequency of days spent in the clear and in the fog", subtitle = "1935-2021", x = "Season", y = "")
  
  
  
# trends

# annual
ff_annual <- season_rh %>% 
  group_by(Class, year) %>% 
  summarise(n_class = n()) %>% 
  group_by(year) %>% 
  mutate(n_total = sum(n_class)) %>% 
  group_by(Class, year) %>% 
  summarise(ff = n_class/n_total, Class = Class) %>% 
  filter(Class == "Fog")

# visual
ggplot(ff_annual, aes(x= year, y = ff)) + geom_point() + geom_smooth(method = "lm", se = F)

ff_sen <- sens.slope(ff_annual$ff, conf.level = 0.95)
View(ff_sen)


# seasonal
```


# Annual trend table 
```{r}
# add decade column --- Nevermind!
# decadal <- season_rh %>% 
#   mutate(year = as.character(year),
#     decade = case_when(
#            year <= 1945 ~ "1",
#            between(year, 1946, 1955)  ~ "2",
#            between(year, 1956, 1965)  ~ "3",
#            between(year, 1966, 1975)  ~ "4",
#            between(year, 1976, 1985)  ~ "5",
#            between(year, 1986, 1995)  ~ "6",
#            between(year, 1996, 2005)  ~ "7",
#            between(year, 2006, 2015)  ~ "8",
#            between(year, 2016, 2021)  ~ "9", #this is not a full decade.... is this an issue?
#            TRUE ~ NA_character_))


## Overall data

# calculate annual trends
annual <- season_rh %>% 
  group_by(year) %>% 
  summarise(avg_rh = mean(RH)) %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Overall") 


## Seasonal

annual_season <- season_rh %>% 
  group_by(year_for_winter, season) %>% 
  summarise(avg_rh = mean(RH)) %>% 
  group_by(season) %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Overall") 



## Clear Seasonal

#overall
annual_clear <- season_rh %>% 
  filter(Class == "Clear") %>% # filter for clear
  group_by(year) %>% 
  summarise(avg_rh = mean(RH)) %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Clear") 
#seasonal
annual_season_clear <- season_rh %>% 
  filter(Class == "Clear") %>% 
  group_by(year_for_winter, season) %>% 
  summarise(avg_rh = mean(RH)) %>% 
  group_by(season) %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Clear") 


#combine tables into one and make it look nice
annual_trends_all <- bind_rows(annual, annual_season, annual_clear, annual_season_clear)  %>% 
  mutate(Type = type, 
         " " = ifelse(is.na(season), "Annual", season), 
         annual_trend = format(round(annual_trend, 3), nsmall = 3), # round trend to 3 decimal places
         "Annual trend" = ifelse(p <= 0.05, str_c(annual_trend, "*"), annual_trend), # add asterisks
         "p-value" = p) %>% 
  dplyr::select(" ", "Annual trend", "p-value") %>% 
  distinct() %>%   # removes rows with repeat values
  
  as.data.frame(html_table()) 
  # add mean RH? for 1941 and for 2021



# Making the table, KableExtra package

annual_trends_all %>%
  kbl(caption = "Annual Trends of Relative Humidity from 1935-2021") %>%
  footnote(general = c("*indicates 95% confidence", "Sen's Slope analysis used to calculate annual relative humidity trends")) %>% 
  kable_classic(full_width = F, html_font = "Cambria") %>% 
  group_rows(group_label = "All", start_row = 1, end_row = 5, hline_after = T) %>% 
  group_rows(group_label = "Clear", start_row = 6, end_row = 10) %>% 
  #collapse_rows(columns = 1, valign = "top") %>%  #not working for some reason, but would be ideal grouping
  add_indent(positions = 1:10, level_of_indent = 2, all_cols = T)



```


```{r}
# all these plots looking at counts, but need to look at averages because the decades are uneven

season_rh %>% 
  filter(RH != 100, year == c(1935, 1945, 1955, 1965, 1975, 1985, 1995, 2005, 2015)) %>% 
    ggplot(aes(x = RH, color = as.factor(year), fill = as.factor(year))) + 
    #geom_histogram(bins = 100) + 
    geom_density(alpha = 0.1) 
    #scale_fill_continuous(type = "gradient")



season_rh %>% 
  filter(RH != 100) %>% 
  ggplot(aes(x = as.factor(year), y = RH)) + geom_violin()

season_rh %>% 
  filter(RH == 100) %>% 
  ggplot(aes(x = as.factor(year), y = RH)) + geom_col()
```

