---
title: "Relative Humidity"
author: "Larz von Huene"
date: '2022-06-06'
output: html_document
---

This rmd calculates relative humidity using WMO annex 4.B and analyzes its trends annually, seasonally, and monthly. The relative humidity is used to determine the presence of fog, and then fog frequency is calculated and analyzed.


#Load Packages
```{r, message=FALSE}
library(tidyverse)
library(lubridate)
library(trend)
library(patchwork)
library(kableExtra)

knitr::opts_chunk$set(echo = TRUE)
```

# Load in Data
```{r, show_col_types = FALSE, message = FALSE}
synoptic <- read_csv("Data/b16 synoptic data 1935-2021.csv") %>% 
  dplyr::select(-c(precipitation, precipitation_trace))
pres_weather <- read_csv("Data/b16 hourly present weather split by code 1935-2021.csv")

# Adding 2022 for synoptic
bulb <- read_csv("Data/2022_synoptic_wet_dry_bulb.csv") %>% 
  mutate(date = as.POSIXct(date, format = "%m/%d/%Y %H:%M", tz = "ETC"))
pressure <- read_csv("Data/2022_RH_pressure.csv") %>%
  dplyr::select(-relative_humidity)  # takes out rh measurement from observatory so we can calculate all rh the same way 

synoptic_2022 <- full_join(bulb, pressure, by = "date") 
  
synoptic <- rbind(synoptic, synoptic_2022) # adding 2022
```


# Clean data
```{r}
#convert temperature into °C
synoptic <- synoptic %>% 
  mutate(dry_bulb = as.numeric(dry_bulb), 
         wet_bulb = as.numeric(wet_bulb), 
         pressure_mb = as.numeric(pressure_mb)) %>% # switch from chr to dbl
  mutate(dry_bulb = (dry_bulb-32)*(5/9),
         wet_bulb = (wet_bulb-32)*(5/9))

# Quality control
synoptic <- synoptic %>% 
  filter(wet_bulb != "NULL" &  pressure_mb != "NULL") %>%  # removes all null values (from wet_bulb, dry_bulb, and pressure_mb)
  filter(wet_bulb <= dry_bulb) %>% # ensures that wet bulb is less than or equal to dry bulb for the sake of the RH equation
  filter(dry_bulb-wet_bulb < 10)  # no gaps larger than 10 degrees between wet and dry)

```



# Calculate Relative Humidity 
Calculations from WMO annex 4.B
Methods for calculation are based on those from Seidel (2007)
```{r}
## Liquid calculations

rh_liquid <- function(t, tw, p){
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p) 
  
  sat_vap_pure = 6.112*(exp((17.62*t)/(243.12+t)))
  sat_vap_pure_from_tw = 6.112*(exp((17.62*tw)/(243.12+tw)))
  
  sat_vap_moist = f_p * sat_vap_pure
  
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw)
  
  return(100*vap_p/sat_vap_moist)
}

liquid <- synoptic %>% 
  filter(dry_bulb >= 0) %>% # Seidel's threshold
  mutate(RH = rh_liquid(dry_bulb, wet_bulb, pressure_mb))


## Ice calculations

rh_ice <- function(t, tw, p) {
  f_p = 1.0016 + (3.15*(10^-6)*p) - (0.074/p)
  
  sat_vap_pure = 6.112*(exp((22.46*t)/(272.62+t)))
  sat_vap_pure_from_tw = 6.112*(exp((22.46*tw)/(272.62+tw)))
  
  sat_vap_moist = f_p * sat_vap_pure
  
  vap_p = sat_vap_pure_from_tw*f_p - 6.53*10^-4*(1+0.000944*tw)*p*(t-tw) # Seidel (2007) used psychometric formula for liquid water for both water and ice
  
  return(100*vap_p/sat_vap_moist)
}

ice <- synoptic %>% 
  filter(dry_bulb < 0) %>% # Seidel's threshold
  mutate(RH = rh_ice(dry_bulb, wet_bulb, pressure_mb))


# Combining
rh <- rbind(liquid, ice)



# Quality control
summary(rh$RH) # 3 NAs

nrow(rh[rh$RH<0,]) # 290 negative rh

rh <- rh %>% 
  na.omit() %>% 
  filter(RH >= 0)


# Export dataset 
write.csv(rh, file = "Data/rh_from_humidity_analysis_rmd.csv")

rh <- rh %>% 
  filter(year(date) >= 1941)
```



# Separate by seasons
```{r}
# seasons by month: so winter is 12 and 1-2, spring is 3-5, summer is 6-8, fall is 9-11
season_rh <- rh %>% 
  na.omit() %>% 
  mutate(year = year(date), 
         month = month(date), 
         year = as.numeric(year), 
         month = as.numeric(month),
         year_for_winter = ifelse(month == 12, year + 1, year), # this is to correct for month 12 being in previous year, so that winter can be grouped together in season_summary (below) 
         season = case_when(
           month == 12 ~ "Winter",
           month == 1 ~ "Winter",
           month == 2 ~ "Winter",
           month == 3 ~ "Spring",
           month == 4 ~ "Spring",
           month == 5 ~ "Spring",
           month == 6 ~ "Summer",
           month == 7 ~ "Summer",
           month == 8 ~ "Summer",
           month == 9 ~ "Fall",
           month == 10 ~ "Fall",
           month == 11 ~ "Fall",
           TRUE ~ NA_character_))



# Getting averages first with daily, then monthly, then seasonal
season_summary <- season_rh %>% 
  group_by(date(date), month, season, year_for_winter) %>% 
  summarize(daily_average_RH = mean(RH)) %>% # first get daily averages
  group_by(month, season, year_for_winter) %>% 
  summarize(monthly_average_RH = mean(daily_average_RH)) %>% # get monthly avg from daily avg
  group_by(season, year_for_winter) %>% 
  summarize(average_RH = mean(monthly_average_RH)) # get seasonal avg from monthly avg
  

# comparing with Seidel
season_summary %>% 
  filter(year_for_winter <= 2004) %>% 
  group_by(season) %>% 
  summarize(avg = mean(average_RH))

# Fewer observations 1935-1939
ggplot(season_rh, aes(x = year, fill = season)) + geom_bar()

season_rh %>% 
  group_by(year) %>% 
  summarize(counts = n())


# Clear day differences in seasons
pdf(file = "Figures/rh_density_plot.pdf", width = 8, height = 5)
season_rh %>% 
  filter(RH < 100) %>% 
  mutate(season = factor(season, c("Winter", "Spring", "Summer", "Fall"))) %>% 
  ggplot(aes(x = RH, fill = season, color = season)) + geom_density(alpha = 0.5) + theme_classic() 
dev.off()
```


# Seasonal comparisons

```{r}
## Linear regression of seasonal data for all season 1935 - 2022, using seasonal averages

# fall
fall_rh <- season_summary %>% 
  filter(season == "Fall") # isolate fall data

fall.lm = lm(average_RH ~ year_for_winter, data = fall_rh)
summary(fall.lm)

# spring
spring_rh <- season_summary %>% 
  filter(season == "Spring") # isolate spring data

spring.lm = lm(average_RH ~ year_for_winter, data = spring_rh)
summary(spring.lm)

# summer
summer_rh <- season_summary %>% 
  filter(season == "Summer") # isolate summer data

summer.lm = lm(average_RH ~ year_for_winter, data = summer_rh)
summary(summer.lm)

# winter
winter_rh <- season_summary %>% 
  filter(season == "Winter") # isolate winter data

winter.lm = lm(average_RH ~ year_for_winter, data = winter_rh)
summary(winter.lm)




## Sen Slopes for seasons

# Fall
fall.sen <- sens.slope(fall_rh$average_RH, conf.level = 0.95)
#View(fall.sen)
fall.sen[["p.value"]]

# Winter
winter.sen <- sens.slope(winter_rh$average_RH, conf.level = 0.95)
#View(winter.sen)
winter.sen[["p.value"]] 

# Spring
spring.sen <- sens.slope(spring_rh$average_RH, conf.level = 0.95)
#View(spring.sen)
spring.sen[["p.value"]] 

# Summer
summer.sen <- sens.slope(summer_rh$average_RH, conf.level = 0.95)
#View(summer.sen)
summer.sen[["p.value"]] 


## VISUALS

# all seasons on one graph
pdf(file = "Figures/humidities_seasons.pdf")
season_summary %>% 
  ggplot(aes(year_for_winter, average_RH, color = season)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  facet_wrap(~season) +
  labs(title = "Average Relative Humidities by Season", subtitle = "1935-2022", x = "Year", y = "Relative Humidity (%)", caption = "Note: Winter graph groups winter months consecutively (e.g. December of 1996, January of 1997, and February of 1997)") 
dev.off()

# spring
pdf(file = "Figures/humidities_spring_1935.pdf", width = 8, height = 5)
spring_rh %>% 
  ggplot(aes(year_for_winter, average_RH)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE, color = "grey") +
  theme_minimal() +
  labs(title = "Average Relative Humidities in Spring", subtitle = "1935-2022", x = "Year", y = "Relative Humidity (%)") +
  annotate("text", x=2015, y=92, label = paste("p-value =", round((spring.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=93, label = paste("Sen's Slope =", round((spring.sen[["estimates"]][["Sen's slope"]]), digits = 3)))
dev.off()


# winter
pdf(file = "Figures/humidities_winter_1935.pdf", width = 8, height = 5)
winter_rh %>% 
  ggplot(aes(year_for_winter, average_RH)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE, color = "grey") +
  theme_minimal() +
  labs(title = "Average Relative Humidities in Winter", subtitle = "1935-2022", x = "Year", y = "Relative Humidity (%)") +
  annotate("text", x=2015, y=90.8, label = paste("p-value =", round((winter.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=92, label = paste("Sen's Slope =", round((winter.sen[["estimates"]][["Sen's slope"]]), digits = 3)))
dev.off()
```

Spring is showing significant trends, but with very small slope

# Monthly comparisons, focusing on November 
```{r}
rh <- rh %>% 
  mutate(month = month(date), year = year(date))

month_rh <- rh %>% 
  group_by(date(date), month, year) %>% # get daily averages first
  summarize(average_RH = mean(RH)) %>% 
  group_by(month, year) %>% # get monthly averages from daily avgs
  summarize(average_RH = mean(average_RH))

# overview of all months
ggplot(month_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm) + facet_wrap(~month)


# RH by month
ggplot(month_rh, aes(as.factor(month), average_RH)) + 
  geom_boxplot() + 
  theme_bw() +
  labs(x = "Month", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidity by Month 1935-2022")

# November sen slopes
nov_rh <- month_rh %>% 
  filter(month == 11) 

ggplot(nov_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm)

nov.lm = lm(average_RH ~ year, data = nov_rh)
summary(nov.lm)


nov.sen <- sens.slope(nov_rh$average_RH, conf.level = 0.95)
#View(nov.sen)
nov.sen[["p.value"]] 

# June
june_rh <- month_rh %>% 
  filter(month == 6) 

ggplot(june_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm)

june.lm = lm(average_RH ~ year, data = june_rh)
summary(june.lm)


# September
sept_rh <- month_rh %>% 
  filter(month == 9) 

ggplot(sept_rh, aes(year, average_RH)) + geom_point() + geom_smooth(method = lm)

sept.lm = lm(average_RH ~ year, data = sept_rh)
summary(sept.lm)


sept.sen <- sens.slope(sept_rh$average_RH, conf.level = 0.95)
#View(sept.sen)
sept.sen[["p.value"]] 


# checking with Seidel 2007
rh %>% 
  filter(year(date) <= 2004) %>% 
  summarize(avg = mean(RH))

rh %>% 
  filter(year(date) <= 2004) %>% 
  filter(month(date) == 6 | month(date) == 7 | month(date) == 8) %>% # for JJA
  summarize(avg = mean(RH))


```

```{r}
# adding names to months
month_rh <- month_rh %>% 
  filter(year >= 1941) %>% # for consistency with wind data
   mutate(month_name = case_when(month == 1 ~ "January", 
                                month == 2 ~ "February",
                                month == 3 ~ "March",
                                month == 4 ~ "April",
                                month == 5 ~ "May",
                                month == 6 ~ "June",
                                month == 7 ~ "July",
                                month == 8 ~ "August",
                                month == 9 ~ "September",
                                month == 10 ~ "October",
                                month == 11 ~ "November",
                                month == 12 ~ "December")) 
month_rh$month_name <- factor(month_rh$month_name, levels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))
  


# RH by month
pdf(file = "Figures/avg_humidity_over_year.pdf", width = 8, height = 5)
ggplot(month_rh, aes(as.factor(month_name), average_RH)) + 
  geom_boxplot(fill = "grey", outlier.shape = 1) + 
  theme_classic() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1)) +
  labs(x = "Month", y = "Relative Humidity (%)") +
  ggtitle("Monthly Average Relative Humidities 1941-2022")
dev.off()
```


# Improving visual

```{r}
# JANUARY
jan_rh <- month_rh %>% 
  filter(month == 1) 

jan <- ggplot(jan_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("January") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

jan.sen <- sens.slope(jan_rh$average_RH, conf.level = 0.95)
#View(jan.sen)
jan.sen[["p.value"]] 






# FEBRUARY
feb_rh <- month_rh %>% 
  filter(month == 2) 

feb <- ggplot(feb_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("February") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

feb.sen <- sens.slope(feb_rh$average_RH, conf.level = 0.95)
#View(feb.sen)
feb.sen[["p.value"]] 







# MARCH
mar_rh <- month_rh %>% 
  filter(month == 3) 

mar <- ggplot(mar_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("March") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()


mar.sen <- sens.slope(mar_rh$average_RH, conf.level = 0.95)
#View(mar.sen)
mar.sen[["p.value"]] 










# APRIL
apr_rh <- month_rh %>% 
  filter(month == 4) 

apr <- ggplot(apr_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("April") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()


apr.sen <- sens.slope(apr_rh$average_RH, conf.level = 0.95)
#View(apr.sen)
apr.sen[["p.value"]] 







# MAY
may_rh <- month_rh %>% 
  filter(month == 5) 

may <- ggplot(may_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("May") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()


may.sen <- sens.slope(may_rh$average_RH, conf.level = 0.95)
#View(may.sen)
may.sen[["p.value"]] 








# JUNE
jun_rh <- month_rh %>% 
  filter(month == 6) 

jun <- ggplot(jun_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("June") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

jun.sen <- sens.slope(jun_rh$average_RH, conf.level = 0.95)
#View(jun.sen)
jun.sen[["p.value"]] 





# JULY
jul_rh <- month_rh %>% 
  filter(month == 7) 

jul <- ggplot(jul_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("July") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

jul.sen <- sens.slope(jul_rh$average_RH, conf.level = 0.95)
#View(jul.sen)
jul.sen[["p.value"]] 








# AUGUST
aug_rh <- month_rh %>% 
  filter(month == 8) 

aug <- ggplot(aug_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("August") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

aug.sen <- sens.slope(aug_rh$average_RH, conf.level = 0.95)
#View(aug.sen)
aug.sen[["p.value"]] 







# SEPTEMEBR
sep_rh <- month_rh %>% 
  filter(month == 9) 

sep <- ggplot(sep_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("September") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

sep.sen <- sens.slope(sep_rh$average_RH, conf.level = 0.95)
#View(sep.sen)
sep.sen[["p.value"]] 









# OCTOBER
oct_rh <- month_rh %>% 
  filter(month == 10) 

oct <- ggplot(oct_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("October") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

oct.sen <- sens.slope(oct_rh$average_RH, conf.level = 0.95)
#View(oct.sen)
oct.sen[["p.value"]] 












# NOVEMBER
nov_rh <- month_rh %>% 
  filter(month == 11) 

nov <- ggplot(nov_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("November") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

nov.sen <- sens.slope(nov_rh$average_RH, conf.level = 0.95)
#View(nov.sen)
nov.sen[["p.value"]] 












# DECEMBER
dec_rh <- month_rh %>% 
  filter(month == 12) 

dec <- ggplot(dec_rh, aes(year, average_RH)) + 
  geom_point() + 
  geom_smooth(method = lm, se = F) + 
  ggtitle("December") + 
  labs (x = "Year", y = "Relative Humidity (%)") +
  theme_minimal()

dec.sen <- sens.slope(dec_rh$average_RH, conf.level = 0.95)
#View(dec.sen)
dec.sen[["p.value"]] 


jan + feb + mar + apr + may + jun + jul + aug + sep + oct + nov + dec
```

# Overall RH over years (not split by month)
```{r}
# getting averages 
year_rh <- season_rh %>% 
  group_by(date(date), month, season, year) %>% 
  summarize(daily_average_RH = mean(RH)) %>% # first get daily averages
  group_by(month, season, year) %>% 
  summarize(monthly_average_RH = mean(daily_average_RH)) %>% # get monthly avg from daily avg
  group_by(year) %>% 
  summarize(avg_rh = mean(monthly_average_RH)) # get yearly avg from monthly avg

# overall average 1941 - 2022
year_rh %>% 
  summarize(avg_rh = mean(avg_rh))

pdf(file = "Figures/rh_annual_1941.pdf", width = 8, height = 5)
year.sen <- sens.slope(year_rh$avg_rh, conf.level = 0.95)
ggplot(year_rh, aes(year, avg_rh)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal() +
  ylim(70, 100) +
  xlim(1941, 2022) + # select only years after 1941 for continuity with wind data
  labs(x = "Year", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidities 1941-2022") + 
  annotate("text", x=2015, y=97, label = paste("p-value =", round((year.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=99, label = paste("Sen's Slope =", round((year.sen[["estimates"]][["Sen's slope"]]), digits = 3)))
dev.off()

# add temp to same graph
rh_all_fig <- ggplot(year_rh, aes(year, avg_rh)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal() +
  ylim(70, 100) +
  xlim(1941, 2022) + # select only years after 1941 for continuity with wind data
  labs(x = "Year", y = "Relative Humidity (%)") +
  ggtitle("Average Relative Humidities 1941-2022") + 
  annotate("text", x=2015, y=97, label = paste("p-value =", round((year.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=99, label = paste("Sen's Slope =", round((year.sen[["estimates"]][["Sen's slope"]]), digits = 3)))

temp <- season_rh %>% 
  group_by(date(date), month, season, year) %>% 
  summarize(daily_average_temp = mean(dry_bulb)) %>% # first get daily averages
  group_by(month, season, year) %>% 
  summarize(monthly_average_temp = mean(daily_average_temp)) %>% # get monthly avg from daily avg
  group_by(year) %>% 
  summarize(avg_temp = mean(monthly_average_temp)) # get yearly avg from monthly avg

temp_fig <- ggplot(temp, aes(year, avg_temp)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey", size = 0.8) +
  theme_minimal()

print(temp_fig/rh_all_fig)
```



# In the clear vs. in the clouds
```{r}
# adding fog and clear class
season_rh <- season_rh %>% 
  mutate(Class = ifelse(RH == 100, "Fog", "Clear")) %>% 
  dplyr::select(date, RH, year, month, year_for_winter, season, Class)

# filter for clear 
clear <- season_rh %>% 
  filter(Class == "Clear") 

# average RH each year
clear_summary <- clear %>% 
  group_by(month, year) %>% 
  summarize(avg_rh = mean(RH)) %>% 
  group_by(year) %>% 
  summarize(avg_rh = mean(avg_rh))

# average in the clear 1941 - 2022 
clear_summary %>% 
  summarize(mean(avg_rh))
# visual: dotplot
pdf(file = "Figures/rh_in_the_clear.pdf", width = 8, height = 5)
clear.sen <- sens.slope(clear_summary$avg_rh, conf.level = 0.95)
clear_summary %>% 
  ggplot(aes(x = year, y = avg_rh)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey") +
  theme_minimal() +
  labs(title = "Average Relative Humidity when in the Clear", subtitle = "1935-2022", x = "Year", y = "Average Relative Humidity (%)") + 
  annotate("text", x=2015, y=75, label = paste("p-value =", round((clear.sen[["p.value"]]), digits = 3))) +
  annotate("text", x=2015, y=76, label = paste("Sen's Slope =", round((clear.sen[["estimates"]][["Sen's slope"]]), digits = 3)))
dev.off()

# analysis (Sen's Slope)
clear.sen <- sens.slope(clear_summary$avg_rh, conf.level = 0.95)
#View(clear.sen)
clear.sen[["p.value"]]
clear.sen[["conf.int"]]



# clear vs all visual
pdf(file = "Figures/rh_plot_by_class.pdf", width = 8, height = 5)
rbind
ggplot(year_rh, aes(x = year, y = avg_rh)) + 
  geom_line(data = year_rh, aes(x = year, y = avg_rh), color = "grey50") + geom_smooth(data = year_rh, aes(x = year, y = avg_rh), method = "lm", se = F, color = "grey70", size = 0.5) + 
  geom_line(data = clear_summary, aes(x = year, y = avg_rh), color = "skyblue3", lty = "longdash") + geom_smooth(data = clear_summary, aes(x = year, y = avg_rh), method = "lm", se = F, color = "skyblue2", size = 0.5) +
  theme_minimal() + 
  labs(x = "Year", y = "Relative Humidity (%)", title = "Relative humidity overall and in the clear", subtitle = "1941-2022" )
dev.off()
```


# Fog frequency
```{r}
# Fog frequency by season: all days 1941-2022
ff_seasonal <- season_rh %>% 
  group_by(season, Class) %>% 
  summarise(n_class = n()) %>% # get number of fog and clear days
  group_by(season) %>% 
  summarise(n_season = sum(n_class), n_class = n_class, Class = Class) %>%  # get total number of days in season
  summarise(ff = n_class/n_season, Class = Class)

#what statistical test will look for significant differences between seasons?

# overall fog frequency
ff_overall <- season_rh %>% 
  group_by(Class) %>% 
  summarise(n_class = n()) %>% 
  mutate(n_total = sum(n_class)) %>% 
  summarise(ff = n_class/n_total, Class = Class)

# visual: seasonal          SHOULD I INCORPORATE OVERALL HERE?
pdf(file = "Figures/ff_season.pdf", width = 8, height = 5)
ggplot(ff_seasonal, aes(x = season, y = ff, fill = Class)) + 
  geom_col() +
  scale_fill_manual(values =c("skyblue", "darkgrey")) +
  theme_minimal() +
  labs(title = "Frequency of days spent in the clear and in the fog", subtitle = "1935-2022", x = "Season", y = "")
dev.off()
  
  
# trends

# annual
ff_annual <- season_rh %>% 
  group_by(Class, year) %>% 
  summarise(n_class = n()) %>% 
  group_by(year) %>% 
  mutate(n_total = sum(n_class)) %>% 
  group_by(Class, year) %>% 
  summarise(ff = n_class/n_total, Class = Class) %>% 
  filter(Class == "Fog")

# visual
ggplot(ff_annual, aes(x= year, y = ff)) + geom_point() + geom_smooth(method = "sen", se = F)

ff_sen <- sens.slope(ff_annual$ff, conf.level = 0.95)


# seasonal
ff_seasonal <- season_rh %>% 
  group_by(Class, year, season) %>% 
  summarise(n_class = n()) %>% 
  group_by(year, season) %>% 
  mutate(n_total = sum(n_class)) %>% 
  group_by(Class, year, season) %>% 
  summarise(ff = n_class/n_total, Class = Class) %>% 
  filter(Class == "Fog")




ff_annual_trend <- ff_annual %>% 
  group_by(year) %>% 
  summarize(annual_trend = sens.slope(ff_annual$ff, conf.level = 0.95)[["estimates"]],
            p = sens.slope(ff_annual$ff, conf.level = 0.95)[["p.value"]]) 
  

ff_seasonal_trend <- ff_seasonal %>% 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>% 
  group_by(year, season) %>% 
  summarise(avg_ff = mean(ff)) %>% 
  group_by(season) %>% 
  summarize(annual_trend = sens.slope(avg_ff, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_ff, conf.level = 0.95)[["p.value"]]) 


# combining for table
ff_annual_trends_all <- bind_rows(ff_annual_trend, ff_seasonal_trend)  %>% 
  mutate(decadal_trend = annual_trend*10*100) %>% #switches trend to decadal percentage
  mutate(" " = ifelse(is.na(season), "Annual", as.character(season)), 
         decadal_trend = format(round(decadal_trend, 3), nsmall = 3), # round trend to 3 decimal places
         "Decadal trend (%)" = ifelse(p <= 0.05, str_c(decadal_trend, "*"), decadal_trend), # add asterisks
         "Decadal trend (%)" = ifelse(p <= 0.1, str_c(decadal_trend, "\u2020"), decadal_trend), # add dagger
         "p-value" = p) %>% 
  dplyr::select(" ", "Decadal trend (%)") %>% 
  distinct() %>%   # removes rows with repeat values
  as.data.frame(html_table()) 


ff_table <- ff_annual_trends_all %>% 
  kbl(caption = "Decadal Trends of Fog Frequency from 1941-2022") %>%
  footnote(general = c("\u2020 indicates 90% confidence", "*indicates 95% confidence", "Sen's Slope analysis used to calculate annual fog frequency trends")) %>% 
  kable_classic(full_width = F, html_font = "Cambria")  

save_kable(ff_table, file = "Figures/ff_table.pdf")

```


# Annual trend table 
```{r}
## Diurnal data
## Overall data
trends_diurnal_rh <- read_csv("Figures/trends_diurnal_rh.csv") %>% 
  dplyr::select(2:6,)


# calculate annual trends
annual <- year_rh %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            #conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Overall") 


## Seasonal

annual_season <- season_summary %>% 
  mutate(avg_rh = average_RH, season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>% 
  group_by(season) %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            #conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Overall") 



## Clear Seasonal

#overall
annual_clear <- clear_summary %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            #conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Clear") 
#seasonal
annual_season_clear <- clear %>% 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>% 
  group_by(date(date), month, season, year_for_winter) %>% 
  summarize(avg_rh = mean(RH)) %>% #daily avg
  group_by(month, season, year_for_winter) %>% 
  summarize(avg_rh = mean(avg_rh)) %>% # monthly avg
  group_by(season, year_for_winter) %>% 
  summarize(avg_rh = mean(avg_rh)) %>%  # seasonal avg
  group_by(season) %>% 
  summarize(annual_trend = sens.slope(avg_rh, conf.level = 0.95)[["estimates"]],
            p = sens.slope(avg_rh, conf.level = 0.95)[["p.value"]],
            #conf_int_95 = sens.slope(avg_rh, conf.level = 0.95)[["conf.int"]],
            type = "Clear") 


#combine tables into one and make it look nice
annual_trends_all <- bind_rows(annual, annual_season, annual_clear, annual_season_clear)  %>% 
  mutate(Type = type, 
         decadal_trend = annual_trend*10,
         " " = ifelse(is.na(season), "Annual", as.character(season)),
         decadal_trend = format(round(decadal_trend, 3), nsmall = 3), # round trend to 3 decimal places
         "Decadal trend (%/decade)" = ifelse(p <= 0.05, str_c(decadal_trend, "*"), decadal_trend), # add asterisks
         "p-value" = p) %>% 
  dplyr::select(Type, " ", "Decadal trend (%/decade)") %>% 
  distinct() %>%   # removes rows with repeat values
  as.data.frame(html_table())


# Making the table, KableExtra package
rh_table <- annual_trends_all %>%
  kbl(caption = "Decadal Trends of Relative Humidity from 1941-2022") %>%
  footnote(general = c("*indicates 95% confidence", "Sen's Slope analysis used to calculate annual relative humidity trends")) %>% 
  kable_classic(full_width = F, html_font = "Cambria") %>% 
  group_rows(group_label = "All", start_row = 1, end_row = 5, hline_after = T) %>% 
  group_rows(group_label = "Clear", start_row = 6, end_row = 10) %>% 
  add_indent(positions = 1:10, level_of_indent = 2, all_cols = T) 

save_kable(rh_table, file = "Figures/humidity_table.pdf")



# Another table, including Day and Night
day <- trends_diurnal_rh %>% 
  filter(tod == "Day") 
night <- trends_diurnal_rh %>% 
  filter(tod == "Night") 
  
a <- bind_cols(bind_rows(annual, annual_season), bind_rows(annual_clear, annual_season_clear), day, night) %>% 
  mutate(season = ifelse(is.na(season...4), "Annual", as.character(season...4)),
         decadal_trend_overall = annual_trend...1*10,
         decadal_trend_clear = annual_trend...5*10,
         decadal_trend_day = trend...10*10,
         decadal_trend_night = trend...15*10,
            decadal_trend_overall = format(round(decadal_trend_overall, 3), nsmall = 3), # round trend to 3 decimal places
            decadal_trend_clear = format(round(decadal_trend_clear, 3), nsmall = 3),
            decadal_trend_day = format(round(decadal_trend_day, 3), nsmall = 3),
            decadal_trend_night = format(round(decadal_trend_night, 3), nsmall = 3),
         decadal_trend_overall = ifelse(p...2 <= 0.05, str_c(decadal_trend_overall, "*"), decadal_trend_overall), # add asterisks
         decadal_trend_clear = ifelse(p...6 <= 0.05, str_c(decadal_trend_clear, "*"), decadal_trend_clear), 
         decadal_trend_day = ifelse(p...11 <= 0.05, str_c(decadal_trend_day, "*"), decadal_trend_day), # add asterisks
         decadal_trend_night = ifelse(p...16 <= 0.05, str_c(decadal_trend_night, "*"), decadal_trend_night)) %>% 
  dplyr::select(season, decadal_trend_overall, decadal_trend_clear, decadal_trend_day, decadal_trend_night)
  

b <- data.frame(stringsAsFactors = FALSE, a)
names(x = b) <- c("Season", "Overall", "Clear", "Day", "Night")


table <- b %>%
  kbl(caption = "Decadal Trends of Relative Humidity (%/Decade) from 1941-2022") %>%
  footnote(general = c("*indicates 95% confidence", "Sen's Slope analysis used to calculate annual relative humidity trends")) %>% 
  kable_classic(full_width = F, html_font = "Cambria")
  
save_kable(table, file = "Figures/humidity_table_diurnal.pdf")

```



