---
title: "Merging Data"
author: "Larz von Huene and Naomi Lubkin"
date: '2022-07-02'
output: html_document
---
This rmd contains the bulk of the analysis for this project. 

There was analysis done with both the 1981-2021 data and 1941-2021 data (see "homogenization.rmd" for more details on how homogenization was done). It includes RH calculations again, and then plots counts of events with high RH, winds, and freezing temperatures. 



# Packages
```{r, message=FALSE}
library(tidyverse)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE)

library(here)
here::here()
library(dplyr)
library(patchwork)
library(gamlss)
library(pscl) #for hurdle model 
```

# data
```{r, message=FALSE}
#b16 hourly present weather, split by code, jan 1935-dec 2021
pres_weather <- read_csv("Data/b16 hourly present weather split by code 1935-2021.csv")

#wind data
hourly_wind_raw <- read_csv("Data/b16-hourly wind 1935-2021.csv")
wind_2022 <- read_csv("Data/2022_wind.csv")

#temp data
hourly_temp <- read_csv("Data/b16-hourly temperature 1935-2021.csv") %>% 
  mutate(temperature = as.numeric(temperature)) 
temp_2022 <- read_csv("Data/2022_temp_hourly.csv")

#rh data
synoptic <- read_csv("Data/b16 synoptic data 1935-2021.csv") %>% 
  dplyr::select(-c(precipitation, precipitation_trace))

bulb <- read_csv("Data/2022_synoptic_wet_dry_bulb.csv") %>%  # NEED TO CONVERT DATE
  mutate(date = as.POSIXct(date, format = "%m/%d/%Y %H:%M"))
pressure <- read_csv("Data/2022_RH_pressure.csv")

synoptic_2022 <- full_join(bulb, pressure, by = "date") %>% 
  dplyr::select(-relative_humidity)  # takes out rh measurement from observatory so we can calculate all rh the same way

```

# Update to add 2022 data (doesn't include weather codes)
```{r}
## WIND

#hourly wind data, 1941-2021
hourly_wind_41on <- hourly_wind_raw %>% 
   filter(year(date) >= 1941) 

#hourly wind data, 2022
hourly_wind_22 <- wind_2022 %>% 
  dplyr::select(date, wind_speed) #same columns as 1941 data for merge

#merging : dataframe with 1941-2022 hourly wind averages
hourly_wind <- rbind(hourly_wind_41on, hourly_wind_22)




## RH

synoptic_2022 <- full_join(bulb, pressure, by = "date") %>% 
  dplyr::select(-relative_humidity)  # takes out rh measurement from observatory so we can calculate all rh the same way 
  
synoptic <- rbind(synoptic, synoptic_2022) # adding 2022




## TEMP

#round up hour
temp_2022 <- temp_2022 %>% 
  mutate(time = round_date(temp_2022$time, unit = "hour"), temperature = f_temp, date = time) %>% 
  dplyr::select(date, temperature) 

hourly_temp <- hourly_temp %>% 
  mutate(date = ceiling_date(hourly_temp$date, unit = "hour"))

hourly_temp <- rbind(hourly_temp, temp_2022)

```

# Clean synoptic data
```{r}
synoptic <- synoptic %>% 
  filter(wet_bulb != "NULL" &  pressure_mb != "NULL") %>%  # removes all null values (from wet_bulb, dry_bulb, and pressure_mb)
  mutate(dry_bulb = as.numeric(dry_bulb), 
         wet_bulb = as.numeric(wet_bulb), 
         pressure_mb = as.numeric(pressure_mb)) %>% # switch from chr to dbl
  filter(wet_bulb <= dry_bulb) %>% # ensures that wet bulb is less than or equal to dry bulb for the sake of the RH equation
  filter(dry_bulb-wet_bulb < 10) %>%  # no gaps larger than 10 degrees between wet and dry)
  filter() # UPDATE: no 20 degress increases between readings
```

# Calculate Relative Humidity 
Calculations from WMO annex 4.B
```{r}
# split data into ice and water (based on Seidel's distinction)
synoptic_rh_liquid <- synoptic %>% 
  filter(wet_bulb >= 32) 
synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

## calculations for water
# saturation vapor pressure
f_p_water <- 1.0016+3.15*(10^-6)*(synoptic_rh_liquid$pressure_mb)-0.074*(synoptic_rh_liquid$pressure_mb)^-1

sat_vap_p_water_pure <- 6.112^(17.62*(synoptic_rh_liquid$dry_bulb)/(243.12+(synoptic_rh_liquid$dry_bulb))) # saturation vapor pressure for pure water
sat_vap_p_water_moist <- f_p_water*sat_vap_p_water_pure  # saturation vapor pressure for moist water
  
# psychometric formulas for the Assmann psychrometer
water_vap_p <- (6.112^(17.62*(synoptic_rh_liquid$wet_bulb)/(243.12+(synoptic_rh_liquid$wet_bulb))))*f_p_water - 6.53*(10^-4)*(1+0.000944*(synoptic_rh_liquid$wet_bulb))*(synoptic_rh_liquid$pressure_mb)*(synoptic_rh_liquid$dry_bulb - synoptic_rh_liquid$wet_bulb)

# Relative humidity
rh_liquid <- 100*water_vap_p/sat_vap_p_water_moist

# liquid tibble
synoptic_rh_liquid <- synoptic_rh_liquid %>% 
  mutate(RH = rh_liquid) 






## calculations for ice

synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

f_p_ice <- 1.0016+(3.15*(10^-6))*(synoptic_rh_ice$pressure_mb)-0.074/synoptic_rh_ice$pressure_mb

sat_vap_p_ice_pure_t <- 6.112*(exp(22.46*(synoptic_rh_ice$dry_bulb)/(272.62+(synoptic_rh_ice$dry_bulb))))
sat_vap_p_ice_moist_t <- sat_vap_p_ice_pure_t*f_p_ice

sat_vap_p_ice_pure_tw <- 6.112*(exp(22.46*(synoptic_rh_ice$wet_bulb)/(272.62+(synoptic_rh_ice$wet_bulb))))
sat_vap_p_ice_moist_tw <- sat_vap_p_ice_pure_tw*f_p_ice

ice_vap_p <- sat_vap_p_ice_pure_t - 5.75*(10^-4)*(1+0.000944*(synoptic_rh_ice$wet_bulb))*(synoptic_rh_ice$pressure_mb)*(synoptic_rh_ice$dry_bulb - synoptic_rh_ice$wet_bulb)

frost_pt <- (272.62*(log(ice_vap_p/6.112*f_p_ice)))/(22.46-(log(ice_vap_p/6.112*f_p_ice)))

sat_vap_p_from_frost <- 6.112*(exp(22.46*(frost_pt)/(272.62+(frost_pt))))
vap_p_from_frost <- sat_vap_p_from_frost - 5.75*(10^-4)*(1+0.000944*(synoptic_rh_ice$wet_bulb))*(synoptic_rh_ice$pressure_mb)*(synoptic_rh_ice$dry_bulb - synoptic_rh_ice$wet_bulb)

rh_ice <- 100*(vap_p_from_frost/sat_vap_p_from_frost)

# ice tibble
synoptic_rh_ice <- synoptic_rh_ice %>% 
  mutate(RH = rh_ice) 
synoptic_rh_ice


## Join ice and liquid into one tibble
rh <- rbind(synoptic_rh_liquid, synoptic_rh_ice) 

# count how many negative values in "rh"
nrow(rh[rh$RH<0,])
  # 73 negative out of 70,874 rows

# percent with negative RH in "rh"
73/70874 # 0.1%

# filter out negative rh
rh <- rh %>% 
  filter(RH >= 0)
```


# fog data stuff (OUTDATED, we are now basing fog off of RH)
```{r}
##FOG: present weather codes (OUTDATED: for final analysis, we used RH fog and not present weather fog)
#sorting through present weather dataframe to select for hours with fog
#"F" means fog was coded for the previous hour, "no_fog" means no fog coded 

hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ "no_fog"))
#"Y" means there was GF (intermittent/patchy fog) coded for the previous hour. 
#no_fake_fog means that GF was NOT coded 
hourly_fog <- hourly_fog %>%
  mutate(fake_fog = case_when (
    pw1 == "GF" ~ "Y", 
    pw2 == "GF" ~ "Y",
    pw3 == "GF" ~ "Y", 
    pw4 == "GF" ~ "Y", 
    pw5 == "GF" ~ "Y", 
    pw6 == "GF" ~ "Y", 
    pw7 == "GF" ~ "Y", 
    pw8 == "GF" ~ "Y", 
    pw9 == "GF" ~ "Y", 
    TRUE ~ "no_fake_fog")) 
#true fog column returns a Y when "F" (fog) is coded for the hour, and "GF" (intermittent/patches of fog) is not
#selecting just for fog and date columns, because those are needed for the merge
hourly_fog_pres <- hourly_fog %>%
  mutate(true_fog = case_when (
    (fog == "F") & (fake_fog == "no_fake_fog") ~ "F", 
    TRUE ~ "NF")) %>%
  dplyr::select(true_fog, date) %>%
  mutate(date = floor_date(hourly_fog$date, unit = "hour"))

#fog dataset with RH to determine fog
#(will go here eventually, need RH's with below freezing)


```


#merge rh with hourly_wind
#for 1981-2022 (merging wind with homogenized 1941-2022 wind data is below)
```{r}
#filter for years of interest (1981-2022)
hourly_wind_81 <- hourly_wind %>% 
  filter(year(date) >= 1981)
#filter for other years of interest (1941-2022)
hourly_wind_41 <- hourly_wind %>% 
  filter(year(date) >= 1941)


#round hourly wind data down to the nearest hour
hourly_wind <- hourly_wind %>% 
  mutate(date = floor_date(hourly_wind$date, unit = "hour"))

# round down hour for fog
hourly_fog_pres <- hourly_fog_pres %>% 
  mutate(date = floor_date(hourly_fog_pres$date, unit = "hour"))

#merge RH and hourly wind data
merged_data <- merge(rh, hourly_wind, by = 'date') %>% 
  dplyr::select(date, dry_bulb, wet_bulb, pressure_mb, RH,  wind_speed)

#and merge in hourly temp data
merged_data <- merge(hourly_temp, merged_data, by = 'date')
#and merge in hourly fog data
merged_data <- merge(hourly_fog_pres, merged_data, by ='date')  
#and select relevant columns
merged_data <- merged_data %>% 
  dplyr::select(date, true_fog, temperature, dry_bulb, wet_bulb, pressure_mb, RH,  wind_speed)

# making wind_speed numeric
merged_data <- merged_data %>% 
  mutate(wind_speed = as.numeric(wind_speed))

#checking temp overlap
ggplot(merged_data, aes(date, temperature)) +
  geom_line() +
  geom_line(aes(x = date, y = dry_bulb), col = "red", alpha = 0.3)
```

#COMPARISON: fog w/ pres weather codes (F and no GF) vs. fog w/ RH (rh>=96%)
```{r}
#analyzing what % of time pres weather codes and RH match

#making a column that returns "F" if RH is 96% or above

fog_comparison <- merged_data %>% 
  mutate(fog_rh = ifelse((RH>=96), "F", "NF")) %>%
  rename(fog_pres = true_fog) %>%
  mutate(match = ifelse(fog_rh == fog_pres, "y_match", "n_match")) %>%
  mutate(date = as.Date(date), 
         month=month(date), 
         year = year(date)) 
fog_comparison$monyear <- str_c(fog_comparison$month, "-", fog_comparison$year)
fog_comparison <- fog_comparison %>%
  dplyr::select(date, match, fog_pres, fog_rh, RH, temperature, dry_bulb, wet_bulb, monyear)

#getting count data for y/n  fog hours (respectively) for each monyear
count_fogs <- fog_comparison %>%
  count(match, monyear) %>%
  filter(match == "y_match")
fog_sums <- aggregate(n ~ monyear, count_fogs, FUN = sum)
fog_proportions <- merge(fog_sums, count_fogs, by = "monyear") %>%
  rename(y_match = n.y, 
         total = n.x) %>%
  separate(monyear, c("month", "year"), sep='-') %>%
  mutate(proportion = y_match/total) %>%
  mutate(year=as.numeric(year))


```


#Plotting proportion of time that fog reported by pres weather vs RH matches 

```{r}
#comparison overall
plot_fogprop_overall <- fog_proportions %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point()

#comparison by month
#first, making dataframes
#nov
prop_nov <- fog_proportions %>%
  filter(month == 11)

#dec
prop_dec <- fog_proportions %>%
  filter(month == 12)

#jan
prop_jan <- fog_proportions %>%
  filter(month == 1)

#feb
prop_feb <- fog_proportions %>%
  filter(month == 2)

#mar
prop_mar <- fog_proportions %>%
  filter(month == 3)

#apr
prop_apr <- fog_proportions %>%
  filter(month == 4)

#may
prop_may <- fog_proportions %>%
  filter(month == 5)

#jun
prop_jun <- fog_proportions %>%
  filter(month == 6)

#and plotting by month
#nov 
plot_prop_nov <- prop_nov %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("November") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion") 

#dec
plot_prop_dec <- prop_dec %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("December") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#jan
plot_prop_jan <- prop_jan %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("January") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#feb
plot_prop_feb <- prop_feb %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("February") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#mar
plot_prop_mar <- prop_mar %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("March") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#apr
plot_prop_apr <- prop_apr %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("April") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#may
plot_prop_may <- prop_may %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("May") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#jun
plot_prop_jun <- prop_jun %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("June") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#combo of months plot
# (plot_prop_nov | plot_prop_dec | plot_prop_jan | plot_prop_feb) / 
#   (plot_prop_mar | plot_prop_apr | plot_prop_may | plot_prop_jun) + plot_annotation(title = "Proportion of times RH fog matches coded fog, by month")
  

```


##1941-2021!
#events of note (freezing, fog, high winds) by month, with RH fog and homogenized wind data 1941-2021 (see homogenization.rmd for more details on homogenization)
```{r}
#clean data
homogenized_hourly <- hourly_wind_41_22 %>% 
  filter(year(date) >= 1941) %>%
  mutate(wind_speed = as.numeric(wind_speed)) 

homogenized_hourly <- homogenized_hourly %>%
  mutate(homogenized_wind = ifelse(year(date) <=1981, homogenized_hourly$wind_speed*1.05, homogenized_hourly$wind_speed)) 

#round hourly wind data down to the nearest hour
homogenized_hourly <- homogenized_hourly %>%
  mutate(date = floor_date(homogenized_hourly$date, unit = "hour"))

#round up hour
hourly_temp <- hourly_temp %>% 
  mutate(date = ceiling_date(hourly_temp$date, unit = "hour"), temperature = as.numeric(temperature))


#merge RH and homogenized hourly wind data
merged_data_41 <- merge(rh, homogenized_hourly, by = 'date')

#and merge in hourly temp data
merged_data_41 <- merge(hourly_temp, merged_data_41, by = 'date')

#and select relevant columns
merged_data_41 <- merged_data_41 %>% 
  dplyr::select(date, temperature, dry_bulb, wet_bulb, pressure_mb, RH,  homogenized_wind)


```

#datasets for each wind threshold, 1941-2022
```{r}
#25mph threshold
#wind_fog column returns a "Y" for synoptic hours that are  below 32 degrees F, have wind averages for the hour 25mph or above, and have 96% or higher relative humidity (indicates fog). returns N if criteria not met
windfog25_41 <- merged_data_41 %>%
  mutate(wind_fog = ifelse(homogenized_wind >= 25  & RH >= 96 & temperature < 32, "Y", "N"))


#60mph threshold
#wind_fog column returns a "Y" for synoptic hours that are  below 32 degrees F, have wind averages for the hour 60mph or above, and have 96% or higher relative humidity (indicates fog). returns N if criteria not met
windfog60_41 <- merged_data_41 %>%
  mutate(wind_fog = ifelse(homogenized_wind >= 60  & RH >= 96 & temperature < 32, "Y", "N"))

#100mph threshold
#wind_fog column returns a "Y" for synoptic hours that are  below 32 degrees F, have wind averages for the hour 100mph or above, and have 96% or higher relative humidity (indicates fog). returns N if criteria not met
windfog100_41 <- merged_data_41 %>%
  mutate(wind_fog = ifelse(homogenized_wind >= 100 & RH >= 96 & temperature < 32, "Y", "N"))
```


#counts of events for each threshold, 1941-2021

```{r}
#events: 25 mph threshold
count_total_25_41 <- windfog25_41 %>% 
  mutate(month = month(date), year = year(date)) %>% 
  mutate(monyear = str_c(month, "-", year)) %>%
  count(monyear, wind_fog) %>% 
  rename(total = n)
  
count_total_events_25_41 <- count_total_25_41 %>%
  count(monyear) %>%
  rename(num_years = n)

counts_25_41 <- merge(count_total_25_41, count_total_events_25_41, by = 'monyear')

counts_final25_41 <- counts_25_41 %>%
  mutate(count_events = ifelse(num_years == 2, total, 0)) %>%
  filter((wind_fog == "Y" & num_years ==2) | (wind_fog == "N" & num_years ==1)) %>%
    separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))


#events: 60 mph threshold
count_total_60_41 <- windfog60_41 %>% 
  mutate(month = month(date), year = year(date)) %>% 
  mutate(monyear = str_c(month, "-", year)) %>%
  count(monyear, wind_fog) %>% 
  rename(total = n)
  
count_total_events_60_41 <- count_total_60_41 %>%
  count(monyear) %>%
  rename(num_years = n)

counts_60_41 <- merge(count_total_60_41, count_total_events_60_41, by = 'monyear')

counts_final60_41 <- counts_60_41 %>%
  mutate(count_events = ifelse(num_years == 2, total, 0)) %>%
  filter((wind_fog == "Y" & num_years ==2) | (wind_fog == "N" & num_years ==1)) %>%
    separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))



#100 mph: getting a column with number of events (or 0)
#counting the number of events where criteria are met (wind_fog column Y's) and not met (wind_fog N's) for each month and year (ex. Jan 1981 = 1-1981 in monyear column)
count_total_100_41 <- windfog100_41 %>% 
  mutate(month = month(date), year = year(date)) %>% 
  mutate(monyear = str_c(month, "-", year)) %>%
  count(monyear, wind_fog) %>% 
  rename(total = n)
  
count_total_events_100_41 <- count_total_100_41 %>%
  count(monyear) %>%
  rename(num_years = n)


counts_100_41 <- merge(count_total_100_41, count_total_events_100_41, by = 'monyear')# %>% 
  #select(monyear, wind_fog, events, total) %>% 
  #mutate(event_count = ifelse(length(monyear) == 2, events, 0))
counts_final100_41 <- counts_100_41 %>%
  mutate(count_events = ifelse(num_years == 2, total, 0)) %>%
  filter((wind_fog == "Y" & num_years ==2) | (wind_fog == "N" & num_years ==1)) %>%
    separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))
```


#making dataframes for each month, overall and for models, 1941-2022 events
```{r}
#25 mph threshold
#first, each month separately, and then each month with just count and year columns for models
#oct 
above25_oct_41 <- counts_final25_41 %>%
  filter(month == 10)
above25oct_formod_41 <- above25_oct_41 %>%
  dplyr::select(count_events, year)


#nov
above25_nov_41 <- counts_final25_41 %>%
  filter(month == 11)
above25nov_formod_41 <- above25_nov_41 %>%
  dplyr::select(count_events, year)

#dec
above25_dec_41 <- counts_final25_41 %>%
  filter(month == 12)
above25dec_formod_41 <- above25_dec_41 %>%
  dplyr::select(count_events, year)

#jan
above25_jan_41 <- counts_final25_41 %>%
  filter(month == 1)
above25jan_formod_41 <- above25_jan_41 %>%
  dplyr::select(count_events, year)

#feb
above25_feb_41 <- counts_final25_41 %>%
  filter(month == 2)
above25feb_formod_41 <- above25_feb_41 %>%
  dplyr::select(count_events, year)

#mar
above25_mar_41 <- counts_final25_41 %>%
  filter(month == 3)
above25mar_formod_41 <- above25_mar_41 %>%
  dplyr::select(count_events, year)

#apr
above25_apr_41 <- counts_final25_41 %>%
  filter(month == 4)
above25apr_formod_41 <- above25_apr_41 %>%
  dplyr::select(count_events, year)

#may
above25_may_41 <- counts_final25_41 %>%
  filter(month == 5)
above25may_formod_41 <- above25_may_41 %>%
  dplyr::select(count_events, year)

#jun
above25_jun_41 <- counts_final25_41 %>%
  filter(month == 6)
above25jun_formod_41 <- above25_jun_41 %>%
  dplyr::select(count_events, year)


# overall
counts_final25_41_sum <- counts_final25_41 %>% 
  group_by(year) %>% 
  summarize(total_events_yr = mean(count_events))
ggplot(counts_final25_41_sum, aes(year, total_events_yr)) +
  geom_point() +
  theme_minimal() +
  geom_smooth(method = "lm", se = F, color = "grey") +
  labs(x = 'Year', y = "Number of Events", title = "Number of Events at 25 mph Threshold", subtitle = "1941-2021")

#sens slope for overall

t25.sen <- sens.slope(counts_final25_41_sum$total_events_yr, conf.level = 0.95)
View(t25.sen)
t25.sen[["p.value"]]


#above 60 mph wind threshold
#first, each month separately, and then each month with just count and year columns for models
#oct 
above60_oct_41 <- counts_final60_41 %>%
  filter(month == 10)
above60oct_formod_41 <- above60_oct_41 %>%
  dplyr::select(count_events, year)


#nov
above60_nov_41 <- counts_final60_41 %>%
  filter(month == 11)
above60nov_formod_41 <- above60_nov_41 %>%
  dplyr::select(count_events, year)

#dec
above60_dec_41 <- counts_final60_41 %>%
  filter(month == 12)
above60dec_formod_41 <- above60_dec_41 %>%
  dplyr::select(count_events, year)

#jan
above60_jan_41 <- counts_final60_41 %>%
  filter(month == 1)
above60jan_formod_41 <- above60_jan_41 %>%
  dplyr::select(count_events, year)

#feb
above60_feb_41 <- counts_final60_41 %>%
  filter(month == 2)
above60feb_formod_41 <- above60_feb_41 %>%
  dplyr::select(count_events, year)

#mar
above60_mar_41 <- counts_final60_41 %>%
  filter(month == 3)
above60mar_formod_41 <- above60_mar_41 %>%
  dplyr::select(count_events, year)

#apr
above60_apr_41 <- counts_final60_41 %>%
  filter(month == 4)
above60apr_formod_41 <- above60_apr_41 %>%
  dplyr::select(count_events, year)

#may
above60_may_41 <- counts_final60_41 %>%
  filter(month == 5)
above60may_formod_41 <- above60_may_41 %>%
  dplyr::select(count_events, year)

#jun
above60_jun_41 <- counts_final60_41 %>%
  filter(month == 6)
above60jun_formod_41 <- above60_jun_41 %>%
  dplyr::select(count_events, year)

# overall
counts_final60_41_sum <- counts_final60_41 %>% 
  group_by(year) %>% 
  summarize(total_events_yr = mean(count_events))
ggplot(counts_final60_41_sum, aes(year, total_events_yr)) +
  geom_point() +
  theme_minimal() +
  labs(x = 'Year', y = "Number of Events", title = "Number of Events at 60 mph Threshold", subtitle = "1941-2021")

#sens slope for overall
t60.sen <- sens.slope(counts_final60_41_sum$total_events_yr, conf.level = 0.95)
View(t60.sen)
t60.sen[["p.value"]]






#above 100 mph wind threshold 
#first, each month separately, and then each month with just count and year columns for models
#oct 
above100_oct_41 <- counts_final100_41 %>%
  filter(month == 10)
above100oct_formod_41 <- above100_oct_41 %>%
  dplyr::select(count_events, year)

#nov
above100_nov_41 <- counts_final100_41 %>%
  filter(month == 11)
above100nov_formod_41 <- above100_nov_41 %>%
  dplyr::select(count_events, year)

#dec
above100_dec_41 <- counts_final100_41 %>%
  filter(month == 12)
above100dec_formod_41 <- above100_dec_41 %>%
  dplyr::select(count_events, year)

#jan
above100_jan_41 <- counts_final100_41 %>%
  filter(month == 1)
above100jan_formod_41 <- above100_jan_41 %>%
  dplyr::select(count_events, year)

#feb
above100_feb_41 <- counts_final100_41 %>%
  filter(month == 2)
above100feb_formod_41 <- above100_feb_41 %>%
  dplyr::select(count_events, year)

#mar
above100_mar_41 <- counts_final100_41 %>%
  filter(month == 3)
above100mar_formod_41 <- above100_mar_41 %>%
  dplyr::select(count_events, year)

#apr
above100_apr_41 <- counts_final100_41 %>%
  filter(month == 4)
above100apr_formod_41 <- above100_apr_41 %>%
  dplyr::select(count_events, year)

#may
above100_may_41 <- counts_final100_41 %>%
  filter(month == 5)
above100may_formod_41 <- above100_may_41 %>%
  dplyr::select(count_events, year)

#jun
above100_jun_41 <- counts_final100_41 %>%
  filter(month == 6)
above100jun_formod_41 <- above100_jun_41 %>%
  dplyr::select(count_events, year)

# overall
counts_final100_41_sum <- counts_final100_41 %>% 
  group_by(year) %>% 
  summarize(total_events_yr = sum(count_events))
ggplot(counts_final100_41_sum, aes(year, total_events_yr)) +
  geom_point() +
  theme_minimal() +
  #scale_y_continuous(breaks = seq(0, 20, by = 1)) +
  labs(x = 'Year', y = "Number of Events", title = "Number of Events at 100 mph Threshold", subtitle = "1941-2021")

#sens slope for overall
t100.sen <- sens.slope(counts_final100_41_sum$total_events_yr, conf.level = 0.95)
View(t100.sen)
t100.sen[["p.value"]]

```

#plots: events 1941-2021
```{r}
#25 mph threshold
#oct
above25_oct_plot_41 <- above25_oct_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("October")

#nov
above25_nov_plot_41 <- above25_nov_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above25_dec_plot_41 <- above25_dec_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above25_jan_plot_41 <- above25_jan_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above25_feb_plot_41 <- above25_feb_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above25_mar_plot_41 <- above25_mar_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above25_apr_plot_41 <- above25_apr_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above25_may_plot_41 <- above25_may_41 %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June
above25_jun_plot_41 <- above25_jun_41 %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#all together 25 mph
allmonths_25_41 <- (above25_nov_plot_41 | above25_dec_plot_41 | above25_jan_plot_41 | above25_feb_plot_41)/ (above25_mar_plot_41 | above25_apr_plot_41 | above25_may_plot_41 | above25_jun_plot_41) 
allmonths_25_41  + plot_annotation(title='Fog, Freezing, and above 25 mph wind hours 1941-2021')







#60mph wind threshold
#oct
above60_oct_plot_41 <- above60_oct_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("October")


#nov
above60_nov_plot_41 <- above60_nov_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above60_dec_plot_41 <- above60_dec_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above60_jan_plot_41 <- above60_jan_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above60_feb_plot_41 <- above60_feb_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above60_mar_plot_41 <- above60_mar_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above60_apr_plot_41 <- above60_apr_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above60_may_plot_41 <- above60_may_41 %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June 
above60_jun_plot_41 <- above60_jun_41 %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#all together 60 mph
allmonths_60_41<- (above60_nov_plot_41 | above60_dec_plot_41 | above60_jan_plot_41| above60_feb_plot_41)/ (above60_mar_plot_41 | above60_apr_plot_41 | above60_may_plot_41 | above60_jun_plot_41) 
allmonths_60_41  + plot_annotation(title='Fog, Freezing, and above 60 mph wind hours 1941-2021')





#100mph wind threshold 
#oct
above100_oct_plot_41 <- above100_oct_41 %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("October")

#nov
above100_nov_plot_41 <- above100_nov_41  %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above100_dec_plot_41  <- above100_dec_41  %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above100_jan_plot_41  <- above100_jan_41  %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above100_feb_plot_41  <- above100_feb_41  %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above100_mar_plot_41  <- above100_mar_41  %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above100_apr_plot_41  <- above100_apr_41  %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above100_may_plot_41  <- above100_may_41  %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June
above100_jun_plot_41  <- above100_jun_41  %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#all together 100 mph
allmonths_100_41 <- (above100_nov_plot_41  | above100_dec_plot_41  | above100_jan_plot_41  | above100_feb_plot_41 )/ (above100_mar_plot_41  | above100_apr_plot_41  | above100_may_plot_41  | above100_jun_plot_41 ) 
allmonths_100_41   + plot_annotation(title='Fog, Freezing, and above 100 mph wind hours 1941-2021')


```

#Dealing with zeros: Hurdle Models
```{r}
oct_hurdle.mod <- hurdle(count_events ~ year, above100oct_formod_41, dist = "poisson", zero.dist = "poisson")
summary(oct_hurdle.mod)


```


#stats: events 1941-2021

```{r}
#25 mph threshold
##Oct
#fitting the distribution 
#Best fit: NBI (AIC:  620.494  )
fitDist(count_events,
        data = above25oct_formod_41, 
        type = "counts", 
        try.gamlss = T)

#model
mod_above25_oct_41 <- gamlss(count_events~year, 
                       family = NBI(), 
                       data=above25oct_formod_41, 
                       control = gamlss.control())
#p-value: 0.338
summary(mod_above25_oct_41)


##Nov
#fitting the distribution 
#Best fit: PIG (AIC:  617.679 )
fitDist(count_events,
        data = above25nov_formod_41, 
        type = "counts", 
        try.gamlss = T)

#model
mod_above25_nov_41 <- gamlss(count_events~year, 
                       family = PIG(), 
                       data=above25nov_formod_41, 
                       control = gamlss.control())
#p-value: 0.93
summary(mod_above25_nov_41)


##Dec
#fitting the distribution 
#Best fit: DPO (AIC: 601.907 )
fitDist(count_events,
        data = above25dec_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_dec_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above25dec_formod_41, 
                       control = gamlss.control())
#p-value: 0.234209    
summary(mod_above25_dec_41)


##Jan
#fitting the distribution 
#Best fit: DPO, AIC: 624.925 
fitDist(count_events,
        data = above25jan_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_jan_41 <- gamlss(count_events~year, 
                       family = DPO(),
                       data=above25jan_formod_41, 
                       control = gamlss.control())
#p-value: 0.317
summary(mod_above25_jan_41)


##Feb
#fitting the distribution 
#Best fit: DPO (AIC: 619.433 )
fitDist(count_events,
        data = above25feb_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_feb_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above25feb_formod_41, 
                       control = gamlss.control(n.cyc = 50))
#p-value: 0.40553
summary(mod_above25_feb_41)


##Mar
#fitting the distribution 
#Best fit: GPO (AIC:  632.089 )
fitDist(count_events,
        data = above25mar_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_mar_41 <- gamlss(count_events~year, 
                       family = GPO(), 
                       data=above25mar_formod_41, 
                       control = gamlss.control(n.cyc = 50))
#p-value: 
summary(mod_above25_mar_41)


##Apr
#fitting the distribution 
#Best fit: DPO (AIC: 635.99)
fitDist(count_events,
        data = above25apr_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_apr_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above25apr_formod_41, 
                       control = gamlss.control())
#p-value: 0.890
summary(mod_above25_apr_41)


##May
#fitting the distribution 
#Best fit: PIG, AIC: 620.32 
(AIC: 635.99)
fitDist(count_events,
        data = above25may_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_may_41 <- gamlss(count_events~year, 
                       family = PIG(), 
                       data=above25may_formod_41, 
                       control = gamlss.control())
#p-value: 0.1725
summary(mod_above25_may_41)


##Jun
#fitting the distribution 
#Best fit: DPO (AIC: 441.133 )
fitDist(count_events,
        data = above25jun_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_jun_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above25jun_formod_41, 
                       control = gamlss.control())
#p-value: 0.669
summary(mod_above25_jun_41)


###############################################################################
#60mph threshold
##Nov
#fitting the distribution 
#Best fit: NBII (AIC: 489.924  )
fitDist(count_events,
        data = above60oct_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_oct_41 <- gamlss(count_events~year, 
                       family = NBII(), 
                       data=above60oct_formod_41, 
                       control = gamlss.control())
#p-value: 0.0644
summary(mod_above60_oct_41)


##Nov
#fitting the distribution 
#Best fit: NBII (AIC: 576.386 )
fitDist(count_events,
        data = above60nov_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_nov_41 <- gamlss(count_events~year, 
                       family = NBII(), 
                       data=above60nov_formod_41, 
                       control = gamlss.control())
#p-value: 0.0162
summary(mod_above60_nov_41)



##Dec
#fitting the distribution 
#Best fit: DPO (AIC: 556.655  )
fitDist(count_events,
        data = above60dec_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_dec_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above60dec_formod_41, 
                       control = gamlss.control())
#p-value: 359
summary(mod_above60_dec_41)



##Jan
#fitting the distribution 
#Best fit: DPO (AIC: 603.473  )
fitDist(count_events,
        data = above60jan_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_jan_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above60jan_formod_41, 
                       control = gamlss.control())
#p-value: 0.685
summary(mod_above60_jan_41)



##Feb
#fitting the distribution 
#Best fit: DPO (AIC: 574.503  )
fitDist(count_events,
        data = above60feb_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_feb_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above60feb_formod_41, 
                       control = gamlss.control())
#p-value: 0.670
summary(mod_above60_feb_41)



##Mar
#fitting the distribution 
#Best fit: DPO (AIC: 602.686  )
fitDist(count_events,
        data = above60mar_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_mar_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above60mar_formod_41, 
                       control = gamlss.control())
#p-value: 0.773
summary(mod_above60_mar_41)


##Apr
#fitting the distribution 
#Best fit: DPO (AIC: .127  )
fitDist(count_events,
        data = above60apr_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_apr_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above60apr_formod_41, 
                       control = gamlss.control())
#p-value: 0.846
summary(mod_above60_apr_41)




##May
#fitting the distribution 
#Best fit: GEOM (AIC: 410.159  )
fitDist(count_events,
        data = above60may_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_may_41 <- gamlss(count_events~year, 
                       family = GEOM(), 
                       data=above60may_formod_41, 
                       control = gamlss.control())
#p-value: 0.669
summary(mod_above60_may_41)





##Jun
#fitting the distribution 
#Best fit: DPO (AIC: 245.968  )
fitDist(count_events,
        data = above60jun_formod_41, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above60_jun_41 <- gamlss(count_events~year, 
                       family = DPO(), 
                       data=above60jun_formod_41, 
                       control = gamlss.control(n.cyc = 100))
#p-value: 0.669
summary(mod_above60_jun_41)

###############################################################################
#because of large number of zeros, not running stats on 100 mph threshold
```

###############################################################################
# #1981-2021!
#looking at events of note (freezing, fog, high winds) by month, with RH instead of present weather codes to determine fog, 1981-2021
## Making datasets for each wind threshold, 1981-2021 
```{r}
#25 mph wind threshold
wind_fog_25 <- merged_data %>% 
 mutate(wind_fog = ifelse(wind_speed >= 25  & RH >= 96 & temperature < 32, "Y", "N"))



#60 mph wind threshold
wind_fog_60 <- merged_data %>% 
  mutate(wind_fog = ifelse(wind_speed >= 60  & RH >= 96 & temperature < 32, "Y", "N"))




#dataframe with Y in column if wind speeds (hourly average) are 100 or above, temp is below 32 (filtering out =32 to avoid edge cases where the hour may be hovering around/above freezing) and RH is 96% or above (based on RH error in Seidel for fog). returns N if criteria not met
wind_fog_100 <- merged_data %>% 
  mutate(wind_fog = ifelse(wind_speed >= 100  & RH >= 96 & temperature < 32, "Y", "N"))



```


```{r}


#events: 25 mph threshold
count_total_25 <- wind_fog_25 %>% 
  mutate(month = month(date), year = year(date)) %>% 
  filter(year > 1980) %>% #weather station change
  mutate(monyear = str_c(month, "-", year)) %>%
  count(monyear, wind_fog) %>% 
  rename(total = n)
  
count_total_events_25 <- count_total_25 %>%
  count(monyear) %>%
  rename(num_years = n)

counts_25 <- merge(count_total_25, count_total_events_25, by = 'monyear')

counts_final25 <- counts_25 %>%
  mutate(count_events = ifelse(num_years == 2, total, 0)) %>%
  filter((wind_fog == "Y" & num_years ==2) | (wind_fog == "N" & num_years ==1)) %>%
    separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))


#events: 60 mph threshold
count_total_60 <- wind_fog_60 %>% 
  mutate(month = month(date), year = year(date)) %>% 
  filter(year > 1980) %>% #weather station change
  mutate(monyear = str_c(month, "-", year)) %>%
  count(monyear, wind_fog) %>% 
  rename(total = n)
  
count_total_events_60 <- count_total_60 %>%
  count(monyear) %>%
  rename(num_years = n)

counts_60 <- merge(count_total_60, count_total_events_60, by = 'monyear')

counts_final60 <- counts_60 %>%
  mutate(count_events = ifelse(num_years == 2, total, 0)) %>%
  filter((wind_fog == "Y" & num_years ==2) | (wind_fog == "N" & num_years ==1)) %>%
    separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))



#100 mph: getting a column with number of events (or 0)
#counting the number of events where criteria are met (wind_fog column Y's) and not met (wind_fog N's) for each month and year (ex. Jan 1981 = 1-1981 in monyear column)
count_total_100 <- wind_fog_100 %>% 
  mutate(month = month(date), year = year(date)) %>% 
  filter(year > 1980) %>% #weather station change
  mutate(monyear = str_c(month, "-", year)) %>%
  count(monyear, wind_fog) %>% 
  rename(total = n)
  
count_total_events_100 <- count_total_100 %>%
  count(monyear) %>%
  rename(num_years = n)


# count_event <- wind_fog_100 %>% 
#   mutate(month = month(date), year = year(date)) %>% 
#   filter(year > 1980) %>% #weather station change
#   count(year, month, wind_fog) %>% 
#   mutate(events = n) %>% 
#   mutate(monyear = str_c(month, "-", year))

counts_100 <- merge(count_total, count_total_events, by = 'monyear')# %>% 
  #select(monyear, wind_fog, events, total) %>% 
  #mutate(event_count = ifelse(length(monyear) == 2, events, 0))
counts_final100 <- counts_100 %>%
  mutate(count_events = ifelse(num_years == 2, total, 0)) %>%
  filter((wind_fog == "Y" & num_years ==2) | (wind_fog == "N" & num_years ==1)) %>%
    separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))
  



# # THIS IS NOT FINISHED: trying to make a proportion of extreme events in the year
# sums <- aggregate(n ~ , counts, FUN = sum)
#   mutate(prop = n(month(date), year(date))/n(month(date), year(date), wind_fog, Y))
# 
# count_fogs <- fog_comparison %>%
#   count(match, monyear) %>%
#   filter(match == "y_match")
# fog_sums <- aggregate(n ~ monyear, count_fogs, FUN = sum)
# fog_proportions <- merge(fog_sums, count_fogs, by = "monyear") %>%
#   rename(y_match = n.y, 
#          total = n.x) %>%
#   separate(monyear, c("month", "year"), sep='-') %>%
#   mutate(proportion = y_match/total) %>%
#   mutate(year=as.numeric(year))

```

Making dataframes for each month, overall and for models
```{r}
#25 mph threshold
#first, each month separately, and then each month with just count and year columns for models
#nov
above25_nov <- counts_final25 %>%
  filter(month == 11)
above25nov_formod <- above25_nov %>%
  dplyr::select(count_events, year)

#dec
above25_dec <- counts_final25 %>%
  filter(month == 12)
above25dec_formod <- above25_dec %>%
  dplyr::select(count_events, year)

#jan
above25_jan <- counts_final25 %>%
  filter(month == 1)
above25jan_formod <- above25_jan %>%
  dplyr::select(count_events, year)

#feb
above25_feb <- counts_final25 %>%
  filter(month == 2)
above25feb_formod <- above25_feb %>%
  dplyr::select(count_events, year)

#mar
above25_mar <- counts_final25 %>%
  filter(month == 3)
above25mar_formod <- above25_mar %>%
  dplyr::select(count_events, year)

#apr
above25_apr <- counts_final25 %>%
  filter(month == 4)
above25apr_formod <- above25_apr %>%
  dplyr::select(count_events, year)

#may
above25_may <- counts_final25 %>%
  filter(month == 5)
above25may_formod <- above25_may %>%
  dplyr::select(count_events, year)

#jun
above25_jun <- counts_final25 %>%
  filter(month == 6)
above25jun_formod <- above25_jun %>%
  dplyr::select(count_events, year)



#above 60 mph wind threshold
#first, each month separately, and then each month with just count and year columns for models
#nov
above60_nov <- counts_final60 %>%
  filter(month == 11)
above60nov_formod <- above60_nov %>%
  dplyr::select(count_events, year)

#dec
above60_dec <- counts_final60 %>%
  filter(month == 12)
above60dec_formod <- above60_dec %>%
  dplyr::select(count_events, year)

#jan
above60_jan <- counts_final60 %>%
  filter(month == 1)
above60jan_formod <- above60_jan %>%
  dplyr::select(count_events, year)

#feb
above60_feb <- counts_final60 %>%
  filter(month == 2)
above60feb_formod <- above60_feb %>%
  dplyr::select(count_events, year)

#mar
above60_mar <- counts_final60 %>%
  filter(month == 3)
above60mar_formod <- above60_mar %>%
  dplyr::select(count_events, year)

#apr
above60_apr <- counts_final60 %>%
  filter(month == 4)
above60apr_formod <- above60_apr %>%
  dplyr::select(count_events, year)

#may
above60_may <- counts_final60 %>%
  filter(month == 5)
above60may_formod <- above60_may %>%
  dplyr::select(count_events, year)

#jun
above60_jun <- counts_final60 %>%
  filter(month == 6)
above60jun_formod <- above60_jun %>%
  dplyr::select(count_events, year)

#above 100 mph wind threshold 
#first, each month separately, and then each month with just count and year columns for models
#nov
above100_nov <- counts_final100 %>%
  filter(month == 11)
above100nov_formod <- above100_nov %>%
  dplyr::select(count_events, year)

#dec
above100_dec <- counts_final100 %>%
  filter(month == 12)
above100dec_formod <- above100_dec %>%
  dplyr::select(count_events, year)

#jan
above100_jan <- counts_final100 %>%
  filter(month == 1)
above100jan_formod <- above100_jan %>%
  dplyr::select(count_events, year)

#feb
above100_feb <- counts_final100 %>%
  filter(month == 2)
above100feb_formod <- above100_feb %>%
  dplyr::select(count_events, year)

#mar
above100_mar <- counts_final100 %>%
  filter(month == 3)
above100mar_formod <- above100_mar %>%
  dplyr::select(count_events, year)

#apr
above100_apr <- counts_final100 %>%
  filter(month == 4)
above100apr_formod <- above100_apr %>%
  dplyr::select(count_events, year)

#may
above100_may <- counts_final100 %>%
  filter(month == 5)
above100may_formod <- above100_may %>%
  dplyr::select(count_events, year)

#jun
above100_jun <- counts_final100 %>%
  filter(month == 6)
above100jun_formod <- above100_jun %>%
  dplyr::select(count_events, year)
```


#plots of events by month
```{r}
#25 mph threshold
#nov
above25_nov_plot <- above25_nov %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above25_dec_plot <- above25_dec %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above25_jan_plot <- above25_jan %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above25_feb_plot <- above25_feb %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above25_mar_plot <- above25_mar %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above25_apr_plot <- above25_apr %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above25_may_plot <- above25_may %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June
above25_jun_plot <- above25_jun %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#all together 25 mph
allmonths_25<- (above25_nov_plot | above25_dec_plot | above25_jan_plot | above25_feb_plot)/ (above25_mar_plot | above25_apr_plot | above25_may_plot | above25_jun_plot) 
allmonths_25  + plot_annotation(title='Fog, Freezing, and above 25 mph wind hours 1981-2021')







#60mph wind threshold
#nov
above60_nov_plot <- above60_nov %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above60_dec_plot <- above60_dec %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above60_jan_plot <- above60_jan %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above60_feb_plot <- above60_feb %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above60_mar_plot <- above60_mar %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above60_apr_plot <- above60_apr %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above60_may_plot <- above60_may %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June 
above60_jun_plot <- above60_jun %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#all together 60 mph
allmonths_60<- (above60_nov_plot | above60_dec_plot | above60_jan_plot | above60_feb_plot)/ (above60_mar_plot | above60_apr_plot | above60_may_plot | above60_jun_plot) 
allmonths_60  + plot_annotation(title='Fog, Freezing, and above 60 mph wind hours 1981-2021')





#100mph wind threshold 
#nov
above100_nov_plot <- above100_nov %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above100_dec_plot <- above100_dec %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above100_jan_plot <- above100_jan %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above100_feb_plot <- above100_feb %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above100_mar_plot <- above100_mar %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above100_apr_plot <- above100_apr %>%
  ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above100_may_plot <- above100_may %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June
above100_jun_plot <- above100_jun %>% 
   ggplot(aes(x=year, y=count_events)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#all together 100 mph
allmonths_100<- (above100_nov_plot | above100_dec_plot | above100_jan_plot | above100_feb_plot)/ (above100_mar_plot | above100_apr_plot | above100_may_plot | above100_jun_plot) 
allmonths_100  + plot_annotation(title='Fog, Freezing, and above 100 mph wind hours 1981-2021')





```

