---
title: "Merging Data"
author: "Larz von Huene and Naomi Lubkin"
date: '2022-07-02'
output: html_document
---
# Packages
```{r}
library(tidyverse)
library(lubridate)

knitr::opts_chunk$set(echo = TRUE)

setwd("//wdc-fileserver/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends")

library(here)
here::here()
library(dplyr)
```

# data
```{r}
#b16 hourly present weather, split by code, jan 1935-dec 2021
pres_weather <- read_csv(here("Data", "b16 hourly present weather split by code 1935-2021 (1).csv"))

#b16 hourly daily wind speed, jan 1935-dec2021
hourly_wind <- read_csv(here("Data", "b16-hourly wind 1935-2021.csv")) 

hourly_temp <- read_csv("//wdc-fileserver.mwo.local/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends/Data/b16-hourly temperature 1935-2021.csv")

synoptic <- read_csv("//wdc-fileserver/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends/Data/b16 synoptic data 1935-2021.csv")

```

#RH calculations (above 32F)
```{r}
# Clean data

synoptic <- synoptic %>% 
  filter(wet_bulb != "NULL" &  pressure_mb != "NULL") %>%  # removes all null values (from wet_bulb, dry_bulb, and pressure_mb)
  mutate(dry_bulb = as.numeric(dry_bulb), 
         wet_bulb = as.numeric(wet_bulb), 
         pressure_mb = as.numeric(pressure_mb), 
         precipitation = as.numeric(precipitation)) %>% # switch from chr to dbl
  filter(wet_bulb <= dry_bulb) # ensures that wet bulb is less than or equal to dry bulb for the sake of the RH equation




# Calculations from WMO annex 4.B

# split data into ice and water (based on Seidel's distinction)
synoptic_rh_liquid <- synoptic %>% 
  filter(wet_bulb >= 32) 
synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

# as of 6.29.22 calculations favor liquid measurements
# saturation vapor pressure
f_p_ice <- 1.0016+3.15*(10^-6)*(synoptic_rh_ice$pressure_mb)-0.074*(synoptic_rh_ice$pressure_mb)^-1 # function of pressure
f_p_water <- 1.0016+3.15*(10^-6)*(synoptic_rh_liquid$pressure_mb)-0.074*(synoptic_rh_liquid$pressure_mb)^-1

sat_vap_p_water_pure <- 6.112^(17.62*(synoptic_rh_liquid$dry_bulb)/(243.12+(synoptic_rh_liquid$dry_bulb))) # saturation vapor pressure for pure water
sat_vap_p_water_moist <- f_p_water*sat_vap_p_water_pure  # saturation vapor pressure for moist water
  
sat_vap_p_ice_pure <- 6.112^(22.46*(synoptic_rh_ice$dry_bulb)/(272.62+(synoptic_rh_ice$dry_bulb)))  # saturation vapor pressure for pure ice
sat_vap_p_ice_moist <- f_p_ice*sat_vap_p_ice_pure  # saturation vapor pressure for moist ice?

# Dew point and Frost point (not needed)


# psychometric formulas for the Assmann psychrometer
water_vap_p <- (6.112^(17.62*(synoptic_rh_liquid$wet_bulb)/(243.12+(synoptic_rh_liquid$wet_bulb))))*f_p_water - 6.53*(10^-4)*(1+0.000944*(synoptic_rh_liquid$wet_bulb))*(synoptic_rh_liquid$pressure_mb)*(synoptic_rh_liquid$dry_bulb - synoptic_rh_liquid$wet_bulb)

# Relative humidity
rh <- 100*water_vap_p/sat_vap_p_water_moist





# liquid tibble
synoptic_rh_liquid <- synoptic_rh_liquid %>% 
  mutate(RH = rh, water_vap_p = water_vap_p) 
synoptic_rh_liquid

```
# fog data stuff
```{r}
#fog dataset with present weather to determine fog
hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ "no_fog"))
hourly_fog <- hourly_fog %>%
  mutate(fake_fog = case_when (
    pw1 == "GF" ~ "Y", 
    pw2 == "GF" ~ "Y",
    pw3 == "GF" ~ "Y", 
    pw4 == "GF" ~ "Y", 
    pw5 == "GF" ~ "Y", 
    pw6 == "GF" ~ "Y", 
    pw7 == "GF" ~ "Y", 
    pw8 == "GF" ~ "Y", 
    pw9 == "GF" ~ "Y", 
    TRUE ~ "no_fake_fog")) 

hourly_fog_pres <- hourly_fog %>%
  mutate(true_fog = case_when (
    (fog == "F") & (fake_fog == "no_fake_fog") ~ "Y", 
    TRUE ~ as.character(NA))) %>%
  dplyr::select(true_fog, date) %>%
  mutate(date = floor_date(hourly_fog$date, unit = "hour"))

#fog dataset with RH to determine fog
#(will go here eventually)


```


#merge synoptic_rh_liquid with hourly_wind
```{r}
hourly_wind <- hourly_wind %>% 
  filter(year(date) >= 1981)

hourly_wind <- hourly_wind %>% 
  mutate(date = floor_date(hourly_wind$date, unit = "hour"))

hourly_temp <- hourly_temp %>% 
  mutate(date = floor_date(hourly_temp$date, unit = "hour"), temperature = as.numeric(temperature))

merged_data <- merge(synoptic_rh_liquid, hourly_wind, by = 'date') %>% 
  dplyr::select(date, dry_bulb, wet_bulb, pressure_mb, precipitation, RH,  wind_speed)

merged_data <- merge(hourly_temp, merged_data, by = 'date')
merged_data <- merge(hourly_fog, merged_data, by ='date')

ggplot(merged_data, aes(date, temperature)) +
  geom_line() +
  geom_line(aes(x = date, y = dry_bulb), col = "red")
```

#looking at events of note (freezing, fog, high winds) by month, with RH instead of present weather codes to determine fog
## Making datasets for each wind threshold, and for temp
```{r}
#22 mph wind threshold
wind_22 <- hourly_wind %>% 
  mutate(windy = case_when (wind_speed >21 ~ "Y", 
                            TRUE ~ as.character(NA)))

#45 mph wind threshold
wind_45 <- hourly_wind %>% 
  mutate(windy = case_when (wind_speed >44 ~ "Y", 
                            TRUE ~ as.character(NA)))

#74 mph wind threshold
wind_74 <- hourly_wind %>% 
  mutate(windy = case_when (wind_speed >73 ~ "Y", 
                            TRUE ~ as.character(NA)))

#100mph wind threshold
wind_100 <- hourly_wind %>% 
  mutate(windy = case_when (wind_speed >99 ~ "Y", 
                            TRUE ~ as.character(NA)))

#temperature dataset, below freezing temps
freezing <- hourly_temp %>% 
  mutate(temp = case_when(temperature < 32 ~ "Y", 
                          TRUE ~ as.character(NA)))

```


## merging datasets for analysis
```{r}
# merging: fog with pres weather
coldfog_pres <- merge(hourly_fog_pres, freezing, by = "date" )
  
#merging: fog with RH
#coldfog_RH <- 
  
# adding a month and year column to both for analysis
```


## getting count data
```{r}
# dataset w pres weather fog

#dataset w RH fog
```


## plotting, by month, datasets w pres weather fog
```{r}
#NOVEMBER
#22 mph
#45 mph
#74 mph
#100 mph

#DECEMBER
#22 mph
#45 mph
#74 mph
#100 mph

#JANUARY
#22 mph
#45 mph
#74 mph
#100 mph

#FEBRUARY
#22 mph
#45 mph
#74 mph
#100 mph

#MARCH
#22 mph
#45 mph
#74 mph
#100 mph

#APRIL
#22 mph
#45 mph
#74 mph
#100 mph

#MAY
#22 mph
#45 mph
#74 mph
#100 mph


```

## plotting, by month, datasets w RH fog
```{r}
#NOVEMBER
#22 mph
#45 mph
#74 mph
#100 mph

#DECEMBER
#22 mph
#45 mph
#74 mph
#100 mph

#JANUARY
#22 mph
#45 mph
#74 mph
#100 mph

#FEBRUARY
#22 mph
#45 mph
#74 mph
#100 mph

#MARCH
#22 mph
#45 mph
#74 mph
#100 mph

#APRIL
#22 mph
#45 mph
#74 mph
#100 mph

#MAY
#22 mph
#45 mph
#74 mph
#100 mph
```

