---
title: "pres_weather_code_analysis"
author: "Larz von Huene and Naomi Lubkin"
date: "2023-05-04"
output: html_document
---


This rmd has the analysis using the pres_weather_code data that we opted to not use because RH was a better method for assessing the presence of fog at each hour.


# Packages
```{r, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(here)
here::here()
library(dplyr)
library(patchwork)
library(gamlss)
```

# Data
```{r, message=FALSE}
#b16 hourly present weather, split by code, jan 1935-dec 2021
pres_weather <- read_csv("Data/b16 hourly present weather split by code 1935-2021.csv")

#rh data
rh <- read_csv("Data/rh_from_humidity_analysis_rmd.csv")

```



# Fog data stuff (OUTDATED, we are now basing fog off of RH)
```{r}
##FOG: present weather codes (OUTDATED: for final analysis, we used RH fog and not present weather fog)
#sorting through present weather dataframe to select for hours with fog
#"F" means fog was coded for the previous hour, "no_fog" means no fog coded 

hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ "no_fog"))
#"Y" means there was GF (intermittent/patchy fog) coded for the previous hour. 
#no_fake_fog means that GF was NOT coded 
hourly_fog <- hourly_fog %>%
  mutate(fake_fog = case_when (
    pw1 == "GF" ~ "Y", 
    pw2 == "GF" ~ "Y",
    pw3 == "GF" ~ "Y", 
    pw4 == "GF" ~ "Y", 
    pw5 == "GF" ~ "Y", 
    pw6 == "GF" ~ "Y", 
    pw7 == "GF" ~ "Y", 
    pw8 == "GF" ~ "Y", 
    pw9 == "GF" ~ "Y", 
    TRUE ~ "no_fake_fog")) 
#true fog column returns a Y when "F" (fog) is coded for the hour, and "GF" (intermittent/patches of fog) is not
#selecting just for fog and date columns, because those are needed for the merge
hourly_fog_pres <- hourly_fog %>%
  mutate(true_fog = case_when (
    (fog == "F") & (fake_fog == "no_fake_fog") ~ "F", 
    TRUE ~ "NF")) %>%
  dplyr::select(true_fog, date) %>%
  mutate(date = floor_date(hourly_fog$date, unit = "hour"))


```

# Merge rh with hourly_wind for 1981-2022 (merging wind with homogenized 1941-2022 wind data is below)
```{r}
#filter for years of interest (1981-2022)
hourly_wind_81 <- hourly_wind %>% 
  filter(year(date) >= 1981)
#filter for other years of interest (1941-2022)
hourly_wind_41 <- hourly_wind %>% 
  filter(year(date) >= 1941)


#round hourly wind data down to the nearest hour
hourly_wind <- hourly_wind %>% 
  mutate(date = floor_date(hourly_wind$date, unit = "hour"))

# round down hour for fog
hourly_fog_pres <- hourly_fog_pres %>% 
  mutate(date = floor_date(hourly_fog_pres$date, unit = "hour"))

#merge RH and hourly wind data
merged_data <- merge(rh, hourly_fog_pres, by = 'date') %>% 
  dplyr::select(date, RH, true_fog)


```

#COMPARISON: fog w/ pres weather codes (F and no GF) vs. fog w/ RH (rh>=96%)
```{r}
#analyzing what % of time pres weather codes and RH match

#making a column that returns "F" if RH is 96% or above

fog_comparison <- merged_data %>% 
  mutate(fog_rh = ifelse((RH>=96), "F", "NF")) %>%
  rename(fog_pres = true_fog) %>%
  mutate(match = ifelse(fog_rh == fog_pres, "y_match", "n_match")) %>%
  mutate(date = as.Date(date), 
         month=month(date), 
         year = year(date)) 
fog_comparison$monyear <- str_c(fog_comparison$month, "-", fog_comparison$year)
fog_comparison <- fog_comparison %>%
  dplyr::select(date, match, fog_pres, fog_rh, RH, monyear)

#getting count data for y/n  fog hours (respectively) for each year
fog_proportions <- fog_comparison %>% 
  mutate(year = year(date)) %>% 
  group_by(year) %>% 
  count(fogs_match = match == "y_match") %>% 
  mutate(total_obs = sum(n)) %>% 
  #filter(fogs_match == FALSE) %>% 
  mutate(proportion = n/total_obs) 
```


#Plotting proportion of time that fog reported by pres weather vs RH matches 

```{r}
#comparison overall (by year)
pdf(files = "Figures/RH_Fog_code_matchup.pdf")
fog_proportions %>%
  ggplot(aes(x=year, y=proportion, fill = fogs_match)) +
  geom_col() 
dev.off()


# Overall
fog_proportions %>% 
  filter(fogs_match == F) %>% 
  ungroup() %>% 
  summarise(not_matching = mean(proportion))

```