---
title: "pres_weather_code_analysis"
author: "Larz von Huene and Naomi Lubkin"
date: "2023-05-04"
output: html_document
---


This rmd has the analysis using the pres_weather_code data that we opted to not use because RH was a better method for assessing the presence of fog at each hour.


# Packages
```{r, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(here)
here::here()
library(dplyr)
library(patchwork)
library(gamlss)
library(pscl) #for hurdle model 
```

# Data
```{r, message=FALSE}
#b16 hourly present weather, split by code, jan 1935-dec 2021
pres_weather <- read_csv("Data/b16 hourly present weather split by code 1935-2021.csv")

#wind data
hourly_wind_raw <- read_csv("Data/b16-hourly wind 1935-2021.csv")


#temp data
hourly_temp <- read_csv("Data/b16-hourly temperature 1935-2021.csv") %>% 
  mutate(temperature = as.numeric(temperature)) 

#rh data
synoptic <- read_csv("Data/b16 synoptic data 1935-2021.csv") %>% 
  dplyr::select(-c(precipitation, precipitation_trace))

```


# Clean synoptic data
```{r}
synoptic <- synoptic %>% 
  filter(wet_bulb != "NULL" &  pressure_mb != "NULL") %>%  # removes all null values (from wet_bulb, dry_bulb, and pressure_mb)
  mutate(dry_bulb = as.numeric(dry_bulb), 
         wet_bulb = as.numeric(wet_bulb), 
         pressure_mb = as.numeric(pressure_mb)) %>% # switch from chr to dbl
  filter(wet_bulb <= dry_bulb) %>% # ensures that wet bulb is less than or equal to dry bulb for the sake of the RH equation
  filter(dry_bulb-wet_bulb < 10) %>%  # no gaps larger than 10 degrees between wet and dry)
  filter() # UPDATE: no 20 degress increases between readings
```

# Calculate Relative Humidity 
Calculations from WMO annex 4.B
```{r}
# split data into ice and water (based on Seidel's distinction)
synoptic_rh_liquid <- synoptic %>% 
  filter(wet_bulb >= 32) 
synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

## calculations for water
# saturation vapor pressure
f_p_water <- 1.0016+3.15*(10^-6)*(synoptic_rh_liquid$pressure_mb)-0.074*(synoptic_rh_liquid$pressure_mb)^-1

sat_vap_p_water_pure <- 6.112^(17.62*(synoptic_rh_liquid$dry_bulb)/(243.12+(synoptic_rh_liquid$dry_bulb))) # saturation vapor pressure for pure water
sat_vap_p_water_moist <- f_p_water*sat_vap_p_water_pure  # saturation vapor pressure for moist water
  
# psychometric formulas for the Assmann psychrometer
water_vap_p <- (6.112^(17.62*(synoptic_rh_liquid$wet_bulb)/(243.12+(synoptic_rh_liquid$wet_bulb))))*f_p_water - 6.53*(10^-4)*(1+0.000944*(synoptic_rh_liquid$wet_bulb))*(synoptic_rh_liquid$pressure_mb)*(synoptic_rh_liquid$dry_bulb - synoptic_rh_liquid$wet_bulb)

# Relative humidity
rh_liquid <- 100*water_vap_p/sat_vap_p_water_moist

# liquid tibble
synoptic_rh_liquid <- synoptic_rh_liquid %>% 
  mutate(RH = rh_liquid) 






## calculations for ice

synoptic_rh_ice <- synoptic %>% 
  filter(wet_bulb < 32)

f_p_ice <- 1.0016+(3.15*(10^-6))*(synoptic_rh_ice$pressure_mb)-0.074/synoptic_rh_ice$pressure_mb

sat_vap_p_ice_pure_t <- 6.112*(exp(22.46*(synoptic_rh_ice$dry_bulb)/(272.62+(synoptic_rh_ice$dry_bulb))))
sat_vap_p_ice_moist_t <- sat_vap_p_ice_pure_t*f_p_ice

sat_vap_p_ice_pure_tw <- 6.112*(exp(22.46*(synoptic_rh_ice$wet_bulb)/(272.62+(synoptic_rh_ice$wet_bulb))))
sat_vap_p_ice_moist_tw <- sat_vap_p_ice_pure_tw*f_p_ice

ice_vap_p <- sat_vap_p_ice_pure_t - 5.75*(10^-4)*(1+0.000944*(synoptic_rh_ice$wet_bulb))*(synoptic_rh_ice$pressure_mb)*(synoptic_rh_ice$dry_bulb - synoptic_rh_ice$wet_bulb)

frost_pt <- (272.62*(log(ice_vap_p/6.112*f_p_ice)))/(22.46-(log(ice_vap_p/6.112*f_p_ice)))

sat_vap_p_from_frost <- 6.112*(exp(22.46*(frost_pt)/(272.62+(frost_pt))))
vap_p_from_frost <- sat_vap_p_from_frost - 5.75*(10^-4)*(1+0.000944*(synoptic_rh_ice$wet_bulb))*(synoptic_rh_ice$pressure_mb)*(synoptic_rh_ice$dry_bulb - synoptic_rh_ice$wet_bulb)

rh_ice <- 100*(vap_p_from_frost/sat_vap_p_from_frost)

# ice tibble
synoptic_rh_ice <- synoptic_rh_ice %>% 
  mutate(RH = rh_ice) 
synoptic_rh_ice


## Join ice and liquid into one tibble
rh <- rbind(synoptic_rh_liquid, synoptic_rh_ice) 

# count how many negative values in "rh"
nrow(rh[rh$RH<0,])
  # 73 negative out of 70,874 rows

# percent with negative RH in "rh"
73/70874 # 0.1%

# filter out negative rh
rh <- rh %>% 
  filter(RH >= 0)
```

# Fog data stuff (OUTDATED, we are now basing fog off of RH)
```{r}
##FOG: present weather codes (OUTDATED: for final analysis, we used RH fog and not present weather fog)
#sorting through present weather dataframe to select for hours with fog
#"F" means fog was coded for the previous hour, "no_fog" means no fog coded 

hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ "no_fog"))
#"Y" means there was GF (intermittent/patchy fog) coded for the previous hour. 
#no_fake_fog means that GF was NOT coded 
hourly_fog <- hourly_fog %>%
  mutate(fake_fog = case_when (
    pw1 == "GF" ~ "Y", 
    pw2 == "GF" ~ "Y",
    pw3 == "GF" ~ "Y", 
    pw4 == "GF" ~ "Y", 
    pw5 == "GF" ~ "Y", 
    pw6 == "GF" ~ "Y", 
    pw7 == "GF" ~ "Y", 
    pw8 == "GF" ~ "Y", 
    pw9 == "GF" ~ "Y", 
    TRUE ~ "no_fake_fog")) 
#true fog column returns a Y when "F" (fog) is coded for the hour, and "GF" (intermittent/patches of fog) is not
#selecting just for fog and date columns, because those are needed for the merge
hourly_fog_pres <- hourly_fog %>%
  mutate(true_fog = case_when (
    (fog == "F") & (fake_fog == "no_fake_fog") ~ "F", 
    TRUE ~ "NF")) %>%
  dplyr::select(true_fog, date) %>%
  mutate(date = floor_date(hourly_fog$date, unit = "hour"))

#fog dataset with RH to determine fog
#(will go here eventually, need RH's with below freezing)


```

# Merge rh with hourly_wind for 1981-2022 (merging wind with homogenized 1941-2022 wind data is below)
```{r}
#filter for years of interest (1981-2022)
hourly_wind_81 <- hourly_wind %>% 
  filter(year(date) >= 1981)
#filter for other years of interest (1941-2022)
hourly_wind_41 <- hourly_wind %>% 
  filter(year(date) >= 1941)


#round hourly wind data down to the nearest hour
hourly_wind <- hourly_wind %>% 
  mutate(date = floor_date(hourly_wind$date, unit = "hour"))

# round down hour for fog
hourly_fog_pres <- hourly_fog_pres %>% 
  mutate(date = floor_date(hourly_fog_pres$date, unit = "hour"))

#merge RH and hourly wind data
merged_data <- merge(rh, hourly_wind, by = 'date') %>% 
  dplyr::select(date, dry_bulb, wet_bulb, pressure_mb, RH,  wind_speed)

#and merge in hourly temp data
merged_data <- merge(hourly_temp, merged_data, by = 'date')
#and merge in hourly fog data
merged_data <- merge(hourly_fog_pres, merged_data, by ='date')  

#and select relevant columns
merged_data <- merged_data %>% 
  dplyr::select(date, true_fog, temperature, dry_bulb, wet_bulb, pressure_mb, RH,  wind_speed)

# making wind_speed numeric
merged_data <- merged_data %>% 
  mutate(wind_speed = as.numeric(wind_speed))

#checking temp overlap
ggplot(merged_data, aes(date, temperature)) +
  geom_line() +
  geom_line(aes(x = date, y = dry_bulb), col = "red", alpha = 0.3)
```

#COMPARISON: fog w/ pres weather codes (F and no GF) vs. fog w/ RH (rh>=96%)
```{r}
#analyzing what % of time pres weather codes and RH match

#making a column that returns "F" if RH is 96% or above

fog_comparison <- merged_data %>% 
  mutate(fog_rh = ifelse((RH>=96), "F", "NF")) %>%
  rename(fog_pres = true_fog) %>%
  mutate(match = ifelse(fog_rh == fog_pres, "y_match", "n_match")) %>%
  mutate(date = as.Date(date), 
         month=month(date), 
         year = year(date)) 
fog_comparison$monyear <- str_c(fog_comparison$month, "-", fog_comparison$year)
fog_comparison <- fog_comparison %>%
  dplyr::select(date, match, fog_pres, fog_rh, RH, temperature, dry_bulb, wet_bulb, monyear)

#getting count data for y/n  fog hours (respectively) for each monyear
count_fogs <- fog_comparison %>%
  count(match, monyear) %>%
  filter(match == "y_match")
fog_sums <- aggregate(n ~ monyear, count_fogs, FUN = sum)
fog_proportions <- merge(fog_sums, count_fogs, by = "monyear") %>%
  rename(y_match = n.y, 
         total = n.x) %>%
  separate(monyear, c("month", "year"), sep='-') %>%
  mutate(proportion = y_match/total) %>%
  mutate(year=as.numeric(year))


```


#Plotting proportion of time that fog reported by pres weather vs RH matches 

```{r}
#comparison overall
plot_fogprop_overall <- fog_proportions %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point()

#comparison by month
#first, making dataframes
#nov
prop_nov <- fog_proportions %>%
  filter(month == 11)

#dec
prop_dec <- fog_proportions %>%
  filter(month == 12)

#jan
prop_jan <- fog_proportions %>%
  filter(month == 1)

#feb
prop_feb <- fog_proportions %>%
  filter(month == 2)

#mar
prop_mar <- fog_proportions %>%
  filter(month == 3)

#apr
prop_apr <- fog_proportions %>%
  filter(month == 4)

#may
prop_may <- fog_proportions %>%
  filter(month == 5)

#jun
prop_jun <- fog_proportions %>%
  filter(month == 6)

#and plotting by month
#nov 
plot_prop_nov <- prop_nov %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("November") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion") 

#dec
plot_prop_dec <- prop_dec %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("December") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#jan
plot_prop_jan <- prop_jan %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("January") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#feb
plot_prop_feb <- prop_feb %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("February") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#mar
plot_prop_mar <- prop_mar %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("March") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#apr
plot_prop_apr <- prop_apr %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("April") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#may
plot_prop_may <- prop_may %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("May") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#jun
plot_prop_jun <- prop_jun %>%
  ggplot(aes(x=year, y=proportion)) +
  geom_point() +
  ggtitle("June") +
  theme(axis.text.x = element_text(angle = 60)) +
  labs(x="Year", y="Proportion")

#combo of months plot
# (plot_prop_nov | plot_prop_dec | plot_prop_jan | plot_prop_feb) / 
#   (plot_prop_mar | plot_prop_apr | plot_prop_may | plot_prop_jun) + plot_annotation(title = "Proportion of times RH fog matches coded fog, by month")
  

```