---
title: "Fog, freezing, above 74mph winds"
author: "Naomi Lubkin"
output: html_document
date: '2022-06-21'
---
#NOTE: an updated analysis was done, this is outdated. See "Merging_data.rmd" for updated analysis and thresholds. 


This is the initial analysis done to try and idenfity "significant" weather events (fog with present weather, freezing, and above 74 mph winds). A;though outdated, a lot of this code was modified for the later analysis so I didn't want to delete the rmd. 

#Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
setwd("//wdc-fileserver/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends")
library(here)
here::here()
library(dplyr)
library(gridExtra)
library(lubridate)
library(gamlss)
```

#Loading in the required data
```{r}
#b16 hourly present weather, split by code, jan 1935-dec 2021
pres_weather <- read_csv(here("Data", "b16 hourly present weather split by code 1935-2021 (1).csv"))

#b16 hourly daily wind speed, jan 1935-dec2021
hourly_wind <- read_csv(here("Data", "b16-hourly wind 1935-2021.csv")) 

hourly_temp <- read_csv("//wdc-fileserver.mwo.local/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends/Data/b16-hourly temperature 1935-2021.csv")
```

#Cleaning Data/ setting up required dataframes 
```{r}
#sorting through present weather dataframe to select for hours with fog
hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ as.character(NA))) %>%
  select(date, fog)


#returning dates that meet all 3 requirements (fog, below freezing, strong winds) per year
xtrm_cond <- merge(hourly_fog, hourly_wind, by = 'date')
xtrm_cond <- merge(xtrm_cond, hourly_temp, by = 'date') 
xtrm_cond <- xtrm_cond %>%
  mutate(xtrm= case_when((fog=="F") & (wind_speed >= 74) & (temperature < 32) ~ "Y", 
         TRUE ~ as.character(NA))) %>%
  mutate(date = as.Date(date), 
         month=month(date), 
         year = year(date), 
         monyear = month + year) %>%
  na.omit() 
#combining year and month for analysis purposes
xtrm_cond$monyear <- str_c(xtrm_cond$month, "-", xtrm_cond$year)

count_xtrm <- xtrm_cond %>% 
  count(year, xtrm)

#counting number of times events with all 3 happen per month / making data workable
xtrm_month <- xtrm_cond %>%
  count(monyear, xtrm) %>%
  separate(monyear, c("month", "year")) %>%
  filter(year>=1981) 
      

#Monthly trends: looking at each month separately 
#Nov
xtrm_nov <- xtrm_month %>%
  filter(month == 11) %>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#Dec
xtrm_dec <- xtrm_month %>%
  filter(month == 12)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#Jan
xtrm_jan <- xtrm_month %>%
  filter(month == 1)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#Feb
xtrm_feb <- xtrm_month %>%
  filter(month == 2)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#Mar
xtrm_mar <- xtrm_month %>%
  filter(month == 3)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#Apr
xtrm_apr <- xtrm_month %>%
  filter(month == 4)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#May
xtrm_may <- xtrm_month %>%
  filter(month == 5)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))
#Jun
xtrm_jun <- xtrm_month %>%
  filter(month == 6)%>%
  mutate(year=as.numeric(year), 
         month=as.numeric(month))


```

#Plots: extreme events (all months combined) by year, 1935-2021 and 1981-2021
```{r}
#bar graph of number of events
xtrm_year_plot_35 <- count_xtrm %>%
  ggplot(aes(x=year, y=n)) +
  geom_point(stat= "identity") +
  labs(x="Year", y="Number of Extreme Weather Events") +
  theme_minimal()

xtrm_year_plot_35

#same thing but with 1981-2021 data
xtrm_81 <- count_xtrm %>%
  filter(year>=1981)

xtrm_year_plot_81 <- xtrm_81 %>%
  ggplot(aes(x=year, y=n)) +
  #geom_bar(stat= "identity") +
  geom_point() +
  geom_smooth( se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Extreme Weather Events") +
  theme_minimal()
```

#Plots: broken down by month, 1981-2021
```{r}
#Number of extreme events broken down by month, 1981-2021
#nov
xtrm_nov_plot <- xtrm_nov %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
xtrm_dec_plot <- xtrm_dec %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
xtrm_jan_plot <- xtrm_jan %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
xtrm_feb_plot <- xtrm_feb %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
xtrm_mar_plot <- xtrm_mar %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
xtrm_apr_plot <- xtrm_apr %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")


#may
xtrm_may_plot <- xtrm_may %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")


#june
xtrm_jun_plot <- xtrm_jun %>%
  ggplot(aes(x=year, y=n)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")


#combined plot of extreme events for nov-apr, by individual month
#grid.by giving me some trouble, maybe try patchwork?
extreme_NovApr_plot <-grid.arrange(xtrm_nov_plot, xtrm_dec_plot, xtrm_jan_plot, xtrm_feb_plot, xtrm_mar_plot, xtrm_apr_plot, 
                                nrow = 2, 
                                top = "Number of Extreme Weather Events by Month, 1981-2021")
extreme_NovApr_plot
```

#Models 
```{r}
#Stats!
#All Years
hist(xtrm_month$n)
fitDist(n,
        data = xtrm_month, 
        type = "counts", 
        try.gamlss = T)
#best fit: ZAMBI (Zero Altered Negative binomial type I) (AIC=3138.92)

histDist(xtrm_month$n, 
         "ZANBI", 
         density = T, 
         main = "Number of Extreme Events compared to Negative bionomial type I", 
         xlab = "Number Extreme")
#not a fantastic fit 

descdist(xtrm_month$n)
#squarely w/i beta, but beta is for continuous values btwn 0-1 so is innappropriate

mod_xtrm_all <- gamlss(n ~ year,
                       family = ZANBI(mu.link = "log", sigma.link = "log", nu.link = "logit"), 
                       data = xtrm_month, 
                       control = gamlss.control())
summary(mod_xtrm_all)

#Just November
#best fit: double poisson (AIC=343.852)
fitDist(n,
        data = xtrm_nov, 
        type = "counts", 
        try.gamlss = T)
mod_xtrm_nov <- gamlss(n~year, 
                       family = DPO(), 
                       data=xtrm_nov, 
                       control = gamlss.control())
summary(mod_xtrm_nov)

#Just December
#best fit: Double poisson (DPO), AIC=376.478
fitDist(n,
        data = xtrm_dec, 
        type = "counts", 
        try.gamlss = T)
mod_xtrm_dec <- gamlss(n~year, 
                       family = DPO(), 
                       data=xtrm_dec, 
                       control = gamlss.control())
summary(mod_xtrm_dec)

#Just January
#best fit: Double poisson (DPO), AIC=404.077
fitDist(n,
        data = xtrm_jan, 
        type = "counts", 
        try.gamlss = T)
mod_xtrm_jan <- gamlss(n~year, 
                       family = DPO(), 
                       data=xtrm_jan, 
                       control = gamlss.control())
summary(mod_xtrm_jan)

#Just February 
#best fit: negative bionomial type II (NBII), AIC=378.301
fitDist(n,
        data = xtrm_feb, 
        type = "counts", 
        try.gamlss = T)
mod_xtrm_feb <- gamlss(n~year, 
                       family = NBII(), 
                       data=xtrm_feb, 
                       control = gamlss.control())
summary(mod_xtrm_feb)

#Just March 
#best fit: Double poisson (DPO), AIC=375.277
fitDist(n,
        data = xtrm_mar, 
        type = "counts", 
        try.gamlss = T)
mod_xtrm_mar <- gamlss(n~year, 
                       family = DPO(), 
                       data=xtrm_mar, 
                       control = gamlss.control())
summary(mod_xtrm_mar)

#Just April
#Best fit Poisson Inverse Gaussian (PIG), AIC= 325.018
fitDist(n,
        data = xtrm_apr, 
        type = "counts", 
        try.gamlss = T)
mod_xtrm_apr <- gamlss(n~year, 
                       family = PIG(), 
                       data=xtrm_apr, 
                       control = gamlss.control())
summary(mod_xtrm_apr)
```

