---
title: "Time Period Comparison"
author: "Naomi Lubkin"
date: '2022-07-02'
output: html_document
---
Comparing wind speed data from two time periods: 1940-1980, when wind instruments
were housed at the old Observatory building, and 1981-2021, when wind instruments
were housed on the top of the Sherman Adams Building tower. 

#setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
setwd("//wdc-fileserver/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends")
library(here)
here::here()
library(lubridate)
library(dplyr)

library(LambertW)
library(climatol)
library(snht)
```


#Loading in data
```{r}
#b16 monthly average wind speed (from hourly averages), jan 1935- dec 2021
monthly_avg <- read_csv(here("Data", "b16-hourly monthly wind average 1935-2021.csv"))

#cleaning data
monthly_avg <- monthly_avg %>%
  mutate(monthly_avg_speed = as.numeric(monthly_avg_speed), 
         monthly_avg_spd = round(monthly_avg_speed, digits = 1), 
         date = as.Date(date), 
         month = month(date), 
         year =year(date)) 

wind_spd <- monthly_avg %>%
  filter(year>=1940) %>% 
  dplyr::select(monthly_avg_spd) %>% 
  rename(wind = monthly_avg_spd)

#yearly averages 
#b16 yearly average wind speed (from hourly averages), jan 1935- dec 2021
yearly_avg <-read_csv(here("Data", "b16-hourly yearly wind average 1935-2021.csv"))
yearly_avg <- yearly_avg %>%
  mutate(year_avg_speed = as.numeric(year_avg_speed), 
         year_avg_speed = round(year_avg_speed, digits = 1), 
         date = as.Date(date), 
         month = month(date), 
         year =year(date))


```

#Creating dataframes for analysis

```{r}
# dataframe with daily averages 1941-2021
all_data_monthly <- monthly_avg %>%
  filter(year >= 1940) %>%
  mutate(era = case_when(year <=1980 ~ 1, 
                         year >= 1981 ~ 2))

all_data_yearly <- yearly_avg %>%
  filter(year >= 1940) %>%
  mutate(era = case_when(year <=1980 ~ 1, 
                         year >= 1981 ~ 2))

```



#t-test
```{r}
#TESTING ASSUMPTIONS
#t-tests assume both groups are normally distributed, have equal variances, so need to test for that first
######################################################################################
#doing a shapiro wilk test to test for normality 
#on all data (1941-2021 daily average wind speeds)
shapiro.test(all_data$monthly_avg_spd)
hist(all_data$monthly_avg_spd)
#p-value = 1.359e-11 (less than 0.05, cannot assume normality )
#tryed various transofmrations, none I tried (log, square, cube) really helped
#I think because the sample size is so big I can violate this assumption? 

#t-test between the two time periods
t.test(monthly_avg_spd ~ era, data=all_data)
#p= 0.0007762

#attempted wilcox test. not sure I need to do one, and results seem funky 
#(p-value =1, when I don't think two eras are that different. examine more closely before using
#ran it with alternative=two sided and got almost 0 p-vale instead of p-value 1. need to do more research
x <- all_data$monthly_avg_spd[all_data$era == 1]
y <- all_data$monthly_avg_spd[all_data$era == 2]
wilcox.test(monthly_avg_spd ~ era, data = all_data)
```

#Attempt at automatic homogenization of climatological series (1940-2021)
```{r}
#temp working directory
wd <- tempdir()
wd0 <- setwd(wd)
write.table(wind_spd, file = "wind_1940-2021.dat", col.names=FALSE, row.names = FALSE)
#first need to convert data from csv into txt file

 # write.table(file = 'wind_1941-2021.dat.txt', sep= ",", col.names = FALSE, row.names = FALSE)



#also need a text file with station coordinates for homogen to run 
X <- c(44.27,44.270833)
Y <- c(71.303611, 71.303333)
Z <- c(1920.85, 1925.422)
# X <- c(44,44)
# Y <- c(71, 71)
# Z <- c(1920, 1925)
CODE <- c(0, 0)
NAME <- c("OBS", "SATOWER")
df <- data.frame(X, Y, Z, CODE, NAME)
write.table(df, file = 'wind_1941-2021.est', sep=" ", col.names = FALSE, row.names = FALSE)
#wd <- tempdir()
#wd0 <- setwd(wd)
read.table(file = 'wind_1941-2021.est')

homogen("wind", 1941, 2021)
```


##Standard normal homogeneity test
```{r}
snht_yearly <- snht(yearly_avg$yearly_avg_speed, 80, robust=F, time=NULL)
```

##Manual homogenization 
#based on equation 5 in Gonzalez-Rouco et al (2001)
#this was done more accurately in "homogenization.rmd"-- here, the number I'm multiplying by is a bit off (reducting by 5% instead of increasing by 5%)
```{r}
#f=qa/qb=33.44/35.28=0.9478, so multiply all values pre-inhomogeneity (1940-1980) by 0.9478
wind_homogenized <- all_data_yearly %>%
  mutate(homogenized_spd = ifelse(era==1, year_avg_speed*0.9478, year_avg_speed))



#plot of homogenized data
homogenized_wind_plot <- wind_homogenized %>%
  ggplot(aes(x=year, y = homogenized_spd)) +
  geom_point()+
  ylim(0,40)

non_homogonized_plot <- wind_homogenized %>%
  ggplot(aes(x=year, y=year_avg_speed)) +
  geom_point() +
  ylim(0,40)
```

#t test on homogenized data!
```{r}
t.test(homogenized_spd ~ era, data=wind_homogenized)
```

