---
title: "Wind Threshold: above 25 mph"
author: "Naomi Lubkin"
output: html_document
date: '2025-06-29'
---
Filtering for events with fog and freezing temperatures, and wind speeds above 25 mph. 
There are separate rmd files for other wind thresholds. 

#Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(lubridate)
library(gamlss)
library(gridExtra)
```

#Loading in the required data
```{r, echo=FALSE}
#b16 hourly present weather, split by code, jan 1935-dec 2021

pres_weather <- read_csv(here("Data", "b16 hourly present weather split by code 1935-2021.csv"))

#b16 hourly daily wind speed, jan 1935-dec2021
hourly_wind <- read_csv(here("Data", "b16-hourly wind 1935-2021.csv"))
hourly_wind <- hourly_wind %>% 
  mutate(wind_speed = as.numeric(wind_speed))

hourly_temp <- read_csv(here("Data", "b16-hourly temperature 1935-2021.csv"))
hourly_temp <- hourly_temp %>% 
  mutate(temperature = as.numeric(temperature))
```

#Cleaning Data/ setting up required dataframes (GENERAL- for use w/ all thresholds)
```{r}
#sorting through present weather dataframe to select for hours with fog
hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ "no_fog"))
hourly_fog <- hourly_fog %>%
  mutate(fake_fog = case_when (
    pw1 == "GF" ~ "Y", 
    pw2 == "GF" ~ "Y",
    pw3 == "GF" ~ "Y", 
    pw4 == "GF" ~ "Y", 
    pw5 == "GF" ~ "Y", 
    pw6 == "GF" ~ "Y", 
    pw7 == "GF" ~ "Y", 
    pw8 == "GF" ~ "Y", 
    pw9 == "GF" ~ "Y", 
    TRUE ~ "no_fake_fog")) 

#true fog column returns a Y when "F" (fog) is coded for the hour, and "GF" (intermittent/patches of fog) is not
hourly_fog <- hourly_fog %>%
  mutate(true_fog = case_when (
    (fog == "F") & (fake_fog == "no_fake_fog") ~ "Y", 
    TRUE ~ as.character(NA)))
  #dplyr::select(date, fog)

# for sorting by date!
#TEST_DATE <- subset(hourly_wind, date > "1983-01-01 00:00:00" & date < "1983-01-31 00:00:00")

#making data workable
xtrm_cond <- merge(hourly_fog, hourly_wind, by = 'date')
xtrm_cond <- merge(xtrm_cond, hourly_temp, by = 'date') 

#making dates and wind speeds/temps correct data type
#separating into month, year column so that I can recombine columns into one with just month & year (so I can sort by that later)
xtrm_cond <- xtrm_cond %>%
  mutate(date = as.Date(date), 
         month=month(date), 
         year = year(date)) %>%
  mutate(wind_speed = as.numeric(wind_speed), 
         temperature = as.numeric(temperature))

#combining year and month for analysis purposes
xtrm_cond$monyear <- str_c(xtrm_cond$month, "-", xtrm_cond$year)


#setting up dataframe with a column that returns Y if fog is present and temperature is below freezing
#USE for all wind thresholds
coldFog<-xtrm_cond %>%
  filter(year>1980) %>%
  mutate(freezefog = case_when((true_fog == "Y") & (temperature < 32) ~ "Y", 
         TRUE ~ as.character(NA))) 
```

#THRESHOLD: above 25 mph (1981-2021)
Setting up dataframe
```{r}
#making a column that returns Y if wind speeds are above 25, temps are below 32, and fog is present
above25 <- coldFog %>% 
  mutate(criteria_met= case_when((wind_speed >24) & (freezefog == "Y") ~ "Y", 
                                 TRUE ~  as.character(NA))) %>%
  dplyr::select(wind_speed, temperature, true_fog, monyear, criteria_met)
#counting events by month, making a column with that count, filtering for years later than 1981
#filtering out all "NA" event counts, or times when 3 requirements not met
above25 <- above25 %>%
  count(monyear, criteria_met) 


#all code below in this chunk is for the purpose of getting months with no extreme events
#to show a 0 in the "event_count" column, and then to filter out NA events for 
#months that do have extreme events (so that either the number of extreme events, 
#or a 0, shows up in the extreme events column, and each month is only represented once)
above25_wrangling <- above25 %>% 
  count(monyear) 
above25_attempt<-merge(above25, above25_wrangling, by= 'monyear')
above25_attempt<- above25_attempt %>%
  rename(num_years = n.y, 
         num_events = n.x) 


above25<- above25_attempt %>% 
 mutate(event_count = ifelse(num_years == 2, above25_attempt$num_events, 0)) %>%
  filter((criteria_met == "Y" & num_years == 2) | (is.na(criteria_met)==TRUE & num_years ==1))%>%
  separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))
```

above 25 mph: sorting by month
```{r}
#Monthly trends: looking at each month separately (Nov-April)
#Nov
above25_nov <- above25 %>%
  filter(month == 11) 

#Dec
above25_dec <- above25 %>%
  filter(month == 12)

#Jan
above25_jan <- above25 %>%
  filter(month == 1)

#Feb
above25_feb <- above25 %>%
  filter(month == 2)

#Mar
above25_mar <- above25 %>%
  filter(month == 3)

#Apr
above25_apr <- above25 %>%
  filter(month == 4)

#May
above25_may <- above25 %>%
  filter(month == 5)

#Jun
above25_jun <- above25 %>%
  filter(month == 6)

```

25 plots: overall and by month
```{r}
#overall 
above25_allmonths <- above25 %>%
  ggplot(aes(x=year, y=event_count)) +
  #geom_bar(stat= "identity") +
  geom_point() +
 # geom_smooth( se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Extreme Weather Events") +
  theme_minimal() 
above25_allmonths
#nov
above25_nov_plot <- above25_nov %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above25_dec_plot <- above25_dec %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above25_jan_plot <- above25_jan %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above25_feb_plot <- above25_feb %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above25_mar_plot <- above25_mar %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above25_apr_plot <- above25_apr %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above25_may_plot <- above25_may %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#jun
above25_jun_plot <- above25_jun %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#combined plot of extreme events for nov-apr, by individual month
above25_NovApr_plot <-grid.arrange(above25_nov_plot, above25_dec_plot, above25_jan_plot, above25_feb_plot, above25_mar_plot, above25_apr_plot, 
                                nrow = 2, 
                                top = "Fog, Freezing, and Above 25mph Events, 1981-2021")

above25_NovJun_plot <- grid.arrange(above25_nov_plot, above25_dec_plot, above25_jan_plot, above25_feb_plot, above25_mar_plot, above25_apr_plot, above25_may_plot, above25_jun_plot,
                                nrow = 2, 
                                top = "Fog, Freezing, and Above 25mph Events, 1981-2021")

```

25 stats
```{r}
###Overall 

###By month

##Nov
#fitting the distribution 
#Best fit: Double Poisson (AIC: 467.021)
fitDist(event_count,
        data = above25_nov, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_nov <- gamlss(event_count~year, 
                       family = DPO(), 
                       data=above25_nov, 
                       control = gamlss.control())
#p-value: 0.946
summary(mod_above25_nov)


##Dec
#fitting the distribution 
#Best fit: Double Poisson (AIC: 459.522)
fitDist(event_count,
        data = above25_dec, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_dec <- gamlss(event_count~year, 
                       family = DPO(), 
                       data=above25_dec, 
                       control = gamlss.control())
#p-value: 0.883
summary(mod_above25_dec)


##Jan
#fitting the distribution 
#Best fit: Double Poisson, AIC: 461.815 
fitDist(event_count,
        data = above25_jan, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_jan <- gamlss(event_count~year, 
                       family = DPO(), 
                       data=above25_jan, 
                       control = gamlss.control())
#p-value: 0.2766
summary(mod_above25_jan)


##Feb
#fitting the distribution 
#Best fit: double poisson (AIC:  460.239 )
fitDist(event_count,
        data = above25_feb, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_feb <- gamlss(event_count~year, 
                       family = DPO(), 
                       data=above25_feb, 
                       control = gamlss.control())
#p-value: 0.243
summary(mod_above25_feb)


##Mar
#fitting the distribution 
#Best fit: Poisson inverse gaussian (AIC:   457.598 )
fitDist(event_count,
        data = above25_mar, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_mar <- gamlss(event_count~year, 
                       family = PIG(), 
                       data=above25_mar, 
                       control = gamlss.control())
#p-value: 0.508
summary(mod_above25_mar)


##Apr
#fitting the distribution 
#Best fit: double poisson (AIC: 450.111 
fitDist(event_count,
        data = above25_apr, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_apr <- gamlss(event_count~year, 
                       family = DPO(), 
                       data=above25_apr, 
                       control = gamlss.control())
#p-value: 0.807
summary(mod_above25_apr)

##May
#fitting the distribution 
#Best fit: PIG (AIC: 441.943
fitDist(event_count,
        data = above25_may, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above25_may <- gamlss(event_count~year, 
                       family =PIG(), 
                       data=above25_may, 
                       control = gamlss.control())
#p-value: 0.807
summary(mod_above25_may)


##June
#fitting the distribution 
#Best fit: DPO (AIC: 354.187
fitDist(event_count,
        data = above25_jun, 
        type = "counts", 
        try.gamlss = T)
#model
#dataframe with just relevant columns so model will run
above_25_jun_formod <- above25_jun %>%
  dplyr::select(year, event_count)
mod_above25_jun <- gamlss(event_count~year, 
                       family =DPO(), 
                       data=above_25_jun_formod, 
                       control = gamlss.control())
#p-value: 0.517
summary(mod_above25_jun)


```


# Just Winter counts (Dec, Jan, Feb): 25 mph threshold
```{r}
# seasons by month: so winter is 12 and 1-2, spring is 3-5, summer is 6-8, fall is 9-11
season_above25 <- above25 %>% 
  mutate(month = as.numeric(month),
         year_for_winter = ifelse(month == 12, year + 1, year), # this is to correct for month 12 being in previous year, so that winter can be grouped together in season_summary (below) 
         season = case_when(
           month == 12 ~ "Winter",
           month == 1 ~ "Winter",
           month == 2 ~ "Winter",
           month == 3 ~ "Spring",
           month == 4 ~ "Spring",
           month == 5 ~ "Spring",
           month == 6 ~ "Summer",
           month == 7 ~ "Summer",
           month == 8 ~ "Summer",
           month == 9 ~ "Fall",
           month == 10 ~ "Fall",
           month == 11 ~ "Fall",
           TRUE ~ NA_character_))


# add up events in winter
winter_above25_summary <- season_above25 %>% 
  group_by(season, year_for_winter) %>% 
  summarize(event_count_total = sum(event_count)) %>% 
  filter(season == "Winter")

# visual
ggplot(winter_above25_summary, aes(x = year_for_winter, y = event_count_total)) +
  geom_point() +
  geom_smooth(method = "lm", se = F, color = "grey") +
  theme_minimal() +
  labs(title = "Winter Events Above 25 mph", subtitle = "1981 - 2021", x = "Year", y = "Number of Events") 

# Sen's slope analysis
winter25.sen <- sens.slope(winter_above25_summary$event_count_total, conf.level = 0.95)
View(winter25.sen)
winter25.sen[["p.value"]]
```