---
title: "wind_threshold_100"
author: "Naomi Lubkin"
output: html_document
date: '2022-06-30'
---
Filtering for events with fog and freezing temperatures, and wind speeds above 100 mph. 
There are separate rmd files for other wind thresholds. 

#Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
setwd("//wdc-fileserver/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends")
library(here)
here::here()
library(dplyr)
library(lubridate)
library(gamlss)
library(gamlss.dist)
library(gridExtra)
```

#Loading in the required data
```{r, echo=FALSE}
#b16 hourly present weather, split by code, jan 1935-dec 2021
pres_weather <- read_csv(here("Data", "b16 hourly present weather split by code 1935-2021 (1).csv"))

#b16 hourly daily wind speed, jan 1935-dec2021
hourly_wind <- read_csv(here("Data", "b16-hourly wind 1935-2021.csv"))
hourly_wind <- hourly_wind %>% 
  mutate(wind_speed = as.numeric(wind_speed))

hourly_temp <- read_csv("//wdc-fileserver.mwo.local/Staff/Summit/Interns/Naomi Lubkin/Wind and Humidity Climate Trends/Data/b16-hourly temperature 1935-2021.csv")
hourly_temp <- hourly_temp %>% 
  mutate(temperature = as.numeric(temperature))
```

#Cleaning Data/ setting up required dataframes (GENERAL- for use w/ all thresholds)
```{r}
#sorting through present weather dataframe to select for hours with fog
hourly_fog <- pres_weather %>%
  mutate(fog= case_when(
    pw1 == "F" ~"F", 
    pw2 == "F" ~"F", 
    pw3 =="F" ~"F", 
    pw4 == "F" ~"F", 
    pw5 == "F" ~"F", 
    pw6 == "F" ~"F", 
    pw7 == "F" ~"F", 
    pw8 == "F" ~"F", 
    pw9 == "F" ~"F", 
    TRUE ~ "no_fog"))
hourly_fog <- hourly_fog %>%
  mutate(fake_fog = case_when (
    pw1 == "GF" ~ "Y", 
    pw2 == "GF" ~ "Y",
    pw3 == "GF" ~ "Y", 
    pw4 == "GF" ~ "Y", 
    pw5 == "GF" ~ "Y", 
    pw6 == "GF" ~ "Y", 
    pw7 == "GF" ~ "Y", 
    pw8 == "GF" ~ "Y", 
    pw9 == "GF" ~ "Y", 
    TRUE ~ "no_fake_fog")) 

#true fog column returns a Y when "F" (fog) is coded for the hour, and "GF" (intermittent/patches of fog) is not
hourly_fog <- hourly_fog %>%
  mutate(true_fog = case_when (
    (fog == "F") & (fake_fog == "no_fake_fog") ~ "Y", 
    TRUE ~ as.character(NA)))
  #dplyr::select(date, fog)

# for sorting by date!
#TEST_DATE <- subset(hourly_wind, date > "1983-01-01 00:00:00" & date < "1983-01-31 00:00:00")

#making data workable
xtrm_cond <- merge(hourly_fog, hourly_wind, by = 'date')
xtrm_cond <- merge(xtrm_cond, hourly_temp, by = 'date') 

#making dates and wind speeds/temps correct data type
#separating into month, year column so that I can recombine columns into one with just month & year (so I can sort by that later)
xtrm_cond <- xtrm_cond %>%
  mutate(date = as.Date(date), 
         month=month(date), 
         year = year(date)) %>%
  mutate(wind_speed = as.numeric(wind_speed), 
         temperature = as.numeric(temperature))

#combining year and month for analysis purposes
xtrm_cond$monyear <- str_c(xtrm_cond$month, "-", xtrm_cond$year)


#setting up dataframe with a column that returns Y if fog is present and temperature is below freezing
#USE for all wind thresholds
coldFog<-xtrm_cond %>%
  filter(year>1980) %>%
  mutate(freezefog = case_when((true_fog == "Y") & (temperature < 32) ~ "Y", 
         TRUE ~ as.character(NA))) 

```

#THRESHOLD: above 100 mph (1981-2021)
Setting up dataframe

```{r}
#making a column that returns Y if wind speeds are above 100, temps are below 32, and fog is present
above100 <- coldFog %>% 
  mutate(criteria_met= case_when((wind_speed >99) & (freezefog == "Y") ~ "Y", 
                                 TRUE ~  as.character(NA))) %>%
  dplyr::select(wind_speed, temperature, true_fog, monyear, criteria_met)
#counting events by month, making a column with that count, filtering for years later than 1981
#filtering out all "NA" event counts, or times when 3 requirements not met
above100 <- above100 %>%
  count(monyear, criteria_met) 


#all code below in this chunk is for the purpose of getting months with no extreme events

#to show a 0 in the "event_count" column, and then to filter out NA events for 
#months that do have extreme events (so that either the number of extreme events, 
#or a 0, shows up in the extreme events column, and each month is only represented once)
above100_wrangling <- above100 %>% 
  count(monyear) 
above100_attempt<-merge(above100, above100_wrangling, by= 'monyear')
above100_attempt<- above100_attempt %>%
  rename(num_years = n.y, 
         num_events = n.x) 


above100<- above100_attempt %>% 
 mutate(event_count = ifelse(num_years == 2, above100_attempt$num_events, 0)) %>%
  filter((criteria_met == "Y" & num_years == 2) | (is.na(criteria_met)==TRUE & num_years ==1))%>%
  separate(monyear, c("month", "year"), sep='-')%>%
  mutate(year = as.numeric(year))

```

above 100 mph: sorting by month
```{r}
#Monthly trends: looking at each month separately (Nov-April)
#Nov


above100_nov <- above100 %>%
  filter(month == 11) 
#Dec
above100_dec <- above100 %>%
  filter(month == 12)

#Jan
above100_jan <- above100 %>%
  filter(month == 1)

#Feb
above100_feb <- above100 %>%
  filter(month == 2)

#Mar
above100_mar <- above100 %>%
  filter(month == 3)

#Apr
above100_apr <- above100 %>%
  filter(month == 4)

#May
above100_may <- above100 %>% 
  filter(month == 5)

#May
above100_may <- above100 %>% 
  filter(month == 5)

#June
above100_jun <- above100 %>%
  filter(month == 6)

#################################
#Dataframes with just event_count and year columns for models
#nov
above100_nov_formod <- above100_nov %>% 
  dplyr::select(event_count, year)

#dec
above100_dec_formod <- above100_dec %>% 
  dplyr::select(event_count, year)

#jan
above100_jan_formod <- above100_jan %>% 
  dplyr::select(event_count, year)


#feb
above100_feb_formod <- above100_feb %>% 
  dplyr::select(event_count, year)

#mar
above100_mar_formod <- above100_mar %>% 
  dplyr::select(event_count, year)

#apr
above100_apr_formod <- above100_apr %>% 
  dplyr::select(event_count, year)

#may
above100_may_formod <- above100_may %>% 
  dplyr::select(event_count, year)

#jun
above100_jun_formod <- above100_jun %>% 
  dplyr::select(event_count, year)
```

100mph plots: overall and by month
```{r}
#overall 
above100_allmonths <- above100 %>%
  ggplot(aes(x=year, y=event_count)) +
  #geom_bar(stat= "identity") +
  geom_point() +
 # geom_smooth( se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Extreme Weather Events") +
  theme_minimal() 
above100_allmonths

#nov
above100_nov_plot <- above100_nov %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("November")


#dec
above100_dec_plot <- above100_dec %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("December")


#jan
above100_jan_plot <- above100_jan %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("January")


#feb
above100_feb_plot <- above100_feb %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("February")


#mar
above100_mar_plot <- above100_mar %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
  labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("March")


#apr
above100_apr_plot <- above100_apr %>%
  ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("April")

#may
above100_may_plot <- above100_may %>% 
   ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  #geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("May")

#June
above100_jun_plot <- above100_jun %>% 
   ggplot(aes(x=year, y=event_count)) +
  geom_point() +
  #geom_smooth(formula = y~x, se =FALSE, method = 'lm', show.legend = FALSE) +
 labs(x="Year", y="Number of Events") +
  theme(axis.text.x = element_text(angle = 60)) +
  ggtitle("June")

#combined plot of extreme events for nov-apr, by individual month
above100_NovJun_plot <-grid.arrange(above100_nov_plot, above100_dec_plot, above100_jan_plot, above100_feb_plot, above100_mar_plot, above100_apr_plot, above100_may_plot,   above100_jun_plot,
                                nrow = 2, 
                                top = "Fog, Freezing, and Above 100mph Events, 1981-2021")

```

100mph stats
```{r}
###Overall 

###By month

##Nov
#fitting the distribution 
#Best fit: Negative binomial type I (AIC:  96.0062 )
fitDist(event_count,
        data = above100_nov_formod, 
        type = "counts", 
        try.gamlss = T)

#model
mod_above100_nov <- gamlss(event_count~year, 
                       family = NBI(), 
                       data=above100_nov_formod, 
                       control = gamlss.control())
#p-value: 2.01e-07
summary(mod_above100_nov)


##Dec
#fitting the distribution 
#Best fit: ZINBI (AIC: 159.813)
fitDist(event_count,
        data = above100_dec_formod, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above100_dec <- gamlss(event_count~year, 
                       family = ZINBI(), 
                       data=above100_dec_formod, 
                       control = gamlss.control())
#p-value: 2e-16
summary(mod_above100_dec)


##Jan
#fitting the distribution 
#Best fit: DPO, AIC: 204.848
fitDist(event_count,
        data = above100_jan_formod, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above100_jan <- gamlss(event_count~year, 
                       family = DPO(),
                       data=above100_jan_formod, 
                       control = gamlss.control())
#p-value: 1.94e-11
summary(mod_above100_jan)


##Feb
#fitting the distribution 
#Best fit: NBI (AIC: 185.385)
fitDist(event_count,
        data = above100_feb_formod, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above100_feb <- gamlss(event_count~year, 
                       family = NBI(), 
                       data=above100_feb_formod, 
                       control = gamlss.control(n.cyc = 50))
#p-value: 0.0887
summary(mod_above100_feb)


##Mar
#fitting the distribution 
#Best fit: NBI (AIC: 123.101)
fitDist(event_count,
        data = above100_mar_formod, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above100_mar <- gamlss(event_count~year, 
                       family = NBI(), 
                       data=above100_mar_formod, 
                       control = gamlss.control())
#p-value: 0.337
summary(mod_above100_mar)


##Apr
#fitting the distribution 
#Best fit: NBI (AIC: 109.333)
fitDist(event_count,
        data = above100_apr_formod, 
        type = "counts", 
        try.gamlss = T)
#model
mod_above100_apr <- gamlss(event_count~year, 
                       family = NBI(), 
                       data=above100_apr_formod, 
                       control = gamlss.control())
#p-value: 0.00782
summary(mod_above100_apr)

#not doing stats for may, june because they have only one event each
```